---
title: "Explore resolution pipeline — genus vs family vs functional type"
date: today
date-format: long
date-modified: last-modified
---

<!--
  Supplementary exploration notebook for the resolution-testing pipeline.
  Compares the genus / family / functional_type branches side-by-side.
  Reads exclusively via targets::tar_read() — no re-execution.

  Pipeline store: Data/targets/project_cz/pipeline_test_resolution
  Active config:  project_cz

  Run the pipeline first:
    Sys.setenv(R_CONFIG_ACTIVE = "project_cz")
    targets::tar_make(
      script = here::here(
        "R/02_Main_analyses/pipeline_test_resolution.R"
      ),
      store  = here::here(
        "Data/targets/project_cz/pipeline_test_resolution"
      )
    )
-->

```{r}
#| label: setup

library(here)

here::i_am(
  "R/03_Supplementary_analyses/_Diagnostics/Explore_resolution_pipeline.qmd"
)

source(
  here::here("R/___setup_project___.R")
)

Sys.setenv(R_CONFIG_ACTIVE = "project_cz")

graphical_options <-
  get_active_config("graphical")

fig_width_inch <-
  purrr::chuck(graphical_options, "width") /
  purrr::chuck(graphical_options, "dpi")

fig_height_inch <-
  purrr::chuck(graphical_options, "height") /
  purrr::chuck(graphical_options, "dpi")

path_res_store <-
  here::here("Data/targets/project_cz/pipeline_test_resolution")

path_traits_store <-
  here::here("Data/targets/traits/pipeline_traits")

# Guard: abort rendering if the store is missing.
if (
  !fs::dir_exists(path_res_store)
) {
  cli::cli_abort(
    c(
      "Resolution pipeline store not found at: {path_res_store}",
      "i" = stringr::str_c(
        "Run `targets::tar_make(",
        "script = 'R/02_Main_analyses/pipeline_test_resolution.R',",
        "store  = 'Data/targets/project_cz/pipeline_test_resolution')`",
        " first."
      )
    )
  )
}

# Convenience vector of the three resolutions.
vec_resolutions <-
  base::c("genus", "family", "functional_type")
```

---

# Section 1 — FT classification source

## 1.1 Classification file

```{r}
#| label: s1-ft-file

path_ft_file <-
  targets::tar_read(
    path_ft_classification,
    store = path_res_store
  )

base::cat(stringr::str_glue(
  "FT classification file used:\n  {path_ft_file}\n"
))
```

## 1.2 Load FT classification

```{r}
#| label: s1-load-ft

if (
  !fs::file_exists(path_ft_file)
) {
  cli::cli_abort(
    c(
      "FT classification file not found: {path_ft_file}",
      "i" = "Run pipeline_traits.R (Segment 6) first."
    )
  )
}

data_ft_classification <-
  qs2::qs_read(path_ft_file)

n_distinct_functional_type <-
  dplyr::n_distinct(dplyr::pull(data_ft_classification, "functional_type"))

base::cat(stringr::str_glue(
  "Rows (taxa):           {base::nrow(data_ft_classification)}\n",
  "Distinct FT groups:    {n_distinct_functional_type}\n"
))
```

## 1.3 N taxa per FT group

```{r}
#| label: s1-taxa-per-ft
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 0.6

data_ft_classification |>
  dplyr::count(.data[["functional_type"]]) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = base::factor(.data[["functional_type"]]),
      y = .data[["n"]]
    )
  ) +
  ggplot2::geom_col(fill = "#5588bb") +
  ggplot2::labs(
    x = "Functional type",
    y = "N taxa"
  ) +
  ggplot2::theme_minimal()
```

## 1.4 Silhouette width distribution per FT

```{r}
#| label: s1-sil-boxplot
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 0.6

data_ft_classification |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = base::factor(.data[["functional_type"]]),
      y = .data[["silhouette_width"]]
    )
  ) +
  ggplot2::geom_boxplot(
    fill = "#5588bb",
    alpha = 0.6,
    outlier.size = 0.8
  ) +
  ggplot2::geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "#cc3333"
  ) +
  ggplot2::labs(
    x = "Functional type",
    y = "Silhouette width"
  ) +
  ggplot2::theme_minimal()
```

## 1.5 FT summary table

```{r}
#| label: s1-ft-summary-table

data_ft_classification |>
  dplyr::group_by(.data[["functional_type"]]) |>
  dplyr::summarise(
    n_taxa = dplyr::n(),
    mean_silhouette_width = base::mean(
      .data[["silhouette_width"]], na.rm = TRUE
    ),
    median_silhouette_width = stats::median(
      .data[["silhouette_width"]],
      na.rm = TRUE
    ),
    min_silhouette_width = base::min(.data[["silhouette_width"]], na.rm = TRUE),
    max_silhouette_width = base::max(.data[["silhouette_width"]], na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data[["functional_type"]]) |>
  tinytable::tt()
```

---

# Section 2 — Community data across resolutions

```{r}
#| label: s2-load

list_community_subset <-
  vec_resolutions |>
  rlang::set_names() |>
  purrr::map(
    .f = ~ {
      target_name <-
        stringr::str_glue("data_community_subset_{.x}")

      targets::tar_read_raw(
        target_name,
        store = path_res_store
      )
    }
  )
```

## 2.1 N taxa, datasets, and samples per resolution

```{r}
#| label: s2-overview-table

list_community_subset |>
  purrr::imap(
    .f = ~ .x |>
      dplyr::summarise(
        n_taxa = dplyr::n_distinct(.data[["taxon"]]),
        n_datasets = dplyr::n_distinct(.data[["dataset_name"]]),
        n_samples = dplyr::n_distinct(
          stringr::str_c(.data[["dataset_name"]], .data[["age"]])
        )
      ) |>
      dplyr::mutate(resolution = .y) |>
      dplyr::relocate("resolution")
  ) |>
  purrr::list_rbind() |>
  tinytable::tt()
```

## 2.2 Taxon count distribution (records per taxon)

```{r}
#| label: s2-taxon-count-dist
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch

list_community_subset |>
  purrr::imap(
    .f = ~ .x |>
      dplyr::count(.data[["taxon"]]) |>
      dplyr::mutate(resolution = .y)
  ) |>
  purrr::list_rbind() |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = .data[["n"]],
      fill = .data[["resolution"]]
    )
  ) +
  ggplot2::geom_histogram(bins = 30, alpha = 0.7) +
  ggplot2::facet_wrap(
    ggplot2::vars(.data[["resolution"]]),
    scales = "free_y"
  ) +
  ggplot2::labs(
    x = "N records per taxon",
    y = "Count",
    fill = "Resolution"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

## 2.3 Top-20 taxa per resolution

```{r}
#| label: s2-top-taxa

list_community_subset |>
  purrr::imap(
    .f = ~ .x |>
      dplyr::count(.data[["taxon"]]) |>
      dplyr::arrange(dplyr::desc(.data[["n"]])) |>
      utils::head(20L) |>
      dplyr::mutate(resolution = .y) |>
      dplyr::relocate("resolution")
  ) |>
  purrr::list_rbind() |>
  tinytable::tt()
```

## 2.4 Community taxa and their FT assignments (functional_type branch)

Shows which community taxa mapped to which FT group.

```{r}
#| label: s2-ft-mapping

data_ft_subset <-
  purrr::chuck(list_community_subset, "functional_type")

# The taxon column in the FT-resolved community holds
# the FT label (e.g. "FT_1"). Join back to ft classification
# to inspect the size of each FT group.
data_ft_dataset_summary <-
  data_ft_subset |>
  dplyr::mutate(
    taxon = base::as.character(.data[["taxon"]])
  ) |>
  dplyr::count(.data[["taxon"]], name = "n_records") |>
  dplyr::left_join(
    data_ft_classification |>
      dplyr::mutate(
        taxon = stringr::str_c(
          "FT_",
          base::as.character(.data[["functional_type"]])
        )
      ) |>
      dplyr::count(.data[["taxon"]], name = "n_taxa_in_ft"),
    by = dplyr::join_by(taxon)
  ) |>
  dplyr::arrange(dplyr::desc(.data[["n_records"]]))

data_ft_dataset_summary |>
  tinytable::tt()
```

---

# Section 3 — Model evaluation across resolutions

```{r}
#| label: s3-load

list_model_evaluation <-
  vec_resolutions |>
  rlang::set_names() |>
  purrr::map(
    .f = ~ {
      target_name <-
        stringr::str_glue("model_evaluation_{.x}")

      tryCatch(
        targets::tar_read_raw(
          target_name,
          store = path_res_store
        ),
        error = function(e) {
          base::message(
            stringr::str_glue(
              "Target '{target_name}' not found. ",
              "Has the pipeline completed for '{.x}'?"
            )
          )
          NULL
        }
      )
    }
  ) |>
  purrr::compact()
```

## 3.1 R² comparison across resolutions

```{r}
#| label: s3-r2-comparison
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 0.6

if (
  base::length(list_model_evaluation) > 0L
) {
  list_model_evaluation |>
    purrr::imap(
      .f = ~ {
        tibble::tibble(
          resolution = .y,
          r2_mcfadden = purrr::pluck(.x, "model", "R2-McFadden"),
          r2_nagelkerke = purrr::pluck(.x, "model", "R2-Nagelkerke")
        )
      }
    ) |>
    purrr::list_rbind() |>
    tidyr::pivot_longer(
      cols = c("r2_mcfadden", "r2_nagelkerke"),
      names_to = "metric",
      values_to = "value"
    ) |>
    dplyr::mutate(
      metric = dplyr::case_match(
        .data[["metric"]],
        "r2_mcfadden" ~ "McFadden R²",
        "r2_nagelkerke" ~ "Nagelkerke R²"
      )
    ) |>
    ggplot2::ggplot(
      ggplot2::aes(
        x = .data[["resolution"]],
        y = .data[["value"]],
        fill = .data[["resolution"]]
      )
    ) +
    ggplot2::geom_col(alpha = 0.7) +
    ggplot2::facet_wrap(ggplot2::vars(.data[["metric"]])) +
    ggplot2::labs(
      x = "Taxonomic resolution",
      y = "R²",
      fill = NULL
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "none")
} else {
  base::message("No model_evaluation targets found — Section 3 skipped.")
}
```

## 3.2 Per-species AUC distribution across resolutions

```{r}
#| label: s3-auc-violin
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch

if (
  base::length(list_model_evaluation) > 0L
) {
  list_model_evaluation |>
    purrr::imap(
      .f = ~ {
        data_species_eval <-
          purrr::pluck(.x, "species")

        if (
          base::is.null(data_species_eval)
        ) {
          return(NULL)
        }

        data_species_eval |>
          dplyr::mutate(resolution = .y)
      }
    ) |>
    purrr::compact() |>
    purrr::list_rbind() |>
    dplyr::filter(!base::is.na(.data[["AUC"]])) |>
    ggplot2::ggplot(
      ggplot2::aes(
        x = .data[["resolution"]],
        y = .data[["AUC"]],
        fill = .data[["resolution"]]
      )
    ) +
    ggplot2::geom_violin(alpha = 0.6, draw_quantiles = 0.5) +
    ggplot2::geom_jitter(
      width = 0.1,
      size = 0.8,
      alpha = 0.4
    ) +
    ggplot2::geom_hline(
      yintercept = 0.5,
      linetype = "dashed",
      colour = "#cc3333"
    ) +
    ggplot2::labs(
      x = "Taxonomic resolution",
      y = "AUC",
      fill = NULL
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "none")
}
```

## 3.3 Per-species accuracy distribution across resolutions

```{r}
#| label: s3-accuracy-violin
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch

if (
  base::length(list_model_evaluation) > 0L
) {
  list_model_evaluation |>
    purrr::imap(
      .f = ~ {
        data_species_eval <-
          purrr::pluck(.x, "species")

        if (
          base::is.null(data_species_eval)
        ) {
          return(NULL)
        }

        data_species_eval |>
          dplyr::mutate(resolution = .y)
      }
    ) |>
    purrr::compact() |>
    purrr::list_rbind() |>
    dplyr::filter(!base::is.na(.data[["Accuracy"]])) |>
    ggplot2::ggplot(
      ggplot2::aes(
        x = .data[["resolution"]],
        y = .data[["Accuracy"]],
        fill = .data[["resolution"]]
      )
    ) +
    ggplot2::geom_violin(alpha = 0.6, draw_quantiles = 0.5) +
    ggplot2::geom_jitter(
      width = 0.1,
      size = 0.8,
      alpha = 0.4
    ) +
    ggplot2::labs(
      x = "Taxonomic resolution",
      y = "Accuracy",
      fill = NULL
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "none")
}
```

## 3.4 Top-10 best and worst performing taxa per resolution

```{r}
#| label: s3-best-worst

if (
  base::length(list_model_evaluation) > 0L
) {
  list_model_evaluation |>
    purrr::imap(
      .f = ~ {
        data_species_eval <-
          purrr::pluck(.x, "species")

        if (
          base::is.null(data_species_eval)
        ) {
          return(NULL)
        }

        data_top_5 <-
          data_species_eval |>
          dplyr::arrange(dplyr::desc(.data[["AUC"]])) |>
          utils::head(5L) |>
          dplyr::mutate(rank = "top 5")

        data_bottom_5 <-
          data_species_eval |>
          dplyr::arrange(.data[["AUC"]]) |>
          utils::head(5L) |>
          dplyr::mutate(rank = "bottom 5")

        dplyr::bind_rows(data_top_5, data_bottom_5) |>
          dplyr::mutate(resolution = .y) |>
          dplyr::select("resolution", "rank", dplyr::everything())
      }
    ) |>
    purrr::compact() |>
    purrr::list_rbind() |>
    tinytable::tt()
}
```

---

# Section 4 — ANOVA variance partitioning across resolutions

```{r}
#| label: s4-load

list_model_anova <-
  vec_resolutions |>
  rlang::set_names() |>
  purrr::map(
    .f = ~ {
      target_name <-
        stringr::str_glue("model_anova_{.x}")

      tryCatch(
        targets::tar_read_raw(
          target_name,
          store = path_res_store
        ),
        error = function(e) {
          base::message(
            stringr::str_glue(
              "Target '{target_name}' not found. ",
              "Has the pipeline completed for '{.x}'?"
            )
          )
          NULL
        }
      )
    }
  ) |>
  purrr::compact()
```

## 4.1 Extract variance fractions per resolution

```{r}
#| label: s4-extract-fractions

if (
  base::length(list_model_anova) > 0L
) {
  list_anova_fractions <-
    list_model_anova |>
    purrr::imap(
      .f = ~ {
        tryCatch(
          extract_anova_fractions(anova_obj = .x) |>
            dplyr::mutate(resolution = .y),
          error = function(e) {
            base::message(
              stringr::str_glue(
                "Error extracting ANOVA fractions for '{.y}': ",
                "{conditionMessage(e)}"
              )
            )
            NULL
          }
        )
      }
    ) |>
    purrr::compact()

  base::message(stringr::str_glue(
    "ANOVA fractions extracted for: ",
    "{stringr::str_c(base::names(list_anova_fractions), collapse = ', ')}"
  ))
} else {
  list_anova_fractions <- base::list()
  base::message("No model_anova targets found — Section 4 skipped.")
}
```

## 4.2 Summary table: mean ± SD per component × resolution

```{r}
#| label: s4-summary-table

if (
  base::length(list_anova_fractions) > 0L
) {
  list_anova_fractions |>
    dplyr::bind_rows() |>
    dplyr::group_by(.data[["resolution"]], .data[["component"]]) |>
    dplyr::summarise(
      mean_r2 = base::mean(.data[["R2_Nagelkerke"]], na.rm = TRUE),
      sd_r2 = stats::sd(.data[["R2_Nagelkerke"]], na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(
      .data[["component"]],
      .data[["resolution"]]
    ) |>
    tinytable::tt()
}
```

## 4.3 R²_Nagelkerke per component × resolution

```{r}
#| label: s4-age-line
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 0.8

if (
  base::length(list_anova_fractions) > 0L
) {
  list_anova_fractions |>
    dplyr::bind_rows() |>
    ggplot2::ggplot(
      ggplot2::aes(
        x = .data[["component"]],
        y = .data[["R2_Nagelkerke"]],
        fill = .data[["resolution"]]
      )
    ) +
    ggplot2::geom_col(
      position = ggplot2::position_dodge(0.8),
      alpha = 0.8
    ) +
    ggplot2::labs(
      x = "Component",
      y = "Nagelkerke R²",
      fill = "Resolution"
    ) +
    ggplot2::theme_minimal()
}
```

## 4.4 Stacked bar: mean variance fractions per resolution

```{r}
#| label: s4-stacked-bar
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 0.6

if (
  base::length(list_anova_fractions) > 0L
) {
  list_anova_fractions |>
    dplyr::bind_rows() |>
    dplyr::group_by(.data[["resolution"]], .data[["component"]]) |>
    dplyr::summarise(
      mean_r2 = base::mean(.data[["R2_Nagelkerke"]], na.rm = TRUE),
      .groups = "drop"
    ) |>
    ggplot2::ggplot(
      ggplot2::aes(
        x = .data[["resolution"]],
        y = .data[["mean_r2"]],
        fill = .data[["component"]]
      )
    ) +
    ggplot2::geom_col(
      position = ggplot2::position_stack(),
      alpha = 0.8
    ) +
    ggplot2::labs(
      x = "Taxonomic resolution",
      y = "Mean variance fraction (R²)",
      fill = "Component"
    ) +
    ggplot2::theme_minimal()
}
```

---

*Notebook generated on `r Sys.Date()` from the resolution pipeline store at `Data/targets/project_cz/pipeline_test_resolution`.*

