---
title: "Explore trait pipeline — step-by-step inspection"
date: today
date-format: long
date-modified: last-modified
---

<!--
  Supplementary exploration notebook for the six-segment trait pipeline.
  Reads every intermediate target from the pipeline store (no re-execution).

  Pipeline store: Data/targets/traits_reference_reference/pipeline_traits_reference
  Active config:  traits

  Run the traits pipeline first:
    targets::tar_make(
      script = here::here("R/02_Main_analyses/pipeline_traits_reference.R"),
      store  = here::here("Data/targets/traits_reference_reference/pipeline_traits_reference")
    )
-->

```{r}
#| label: setup

library(here)

here::i_am(
  "R/03_Supplementary_analyses/_Diagnostics/Explore_trait_pipeline.qmd"
)

source(
  here::here("R/___setup_project___.R")
)

Sys.setenv(R_CONFIG_ACTIVE = "project_traits_reference")

graphical_options <-
  get_active_config("graphical")

fig_width_inch <-
  purrr::chuck(graphical_options, "width") /
  purrr::chuck(graphical_options, "dpi")

fig_height_inch <-
  purrr::chuck(graphical_options, "height") /
  purrr::chuck(graphical_options, "dpi")

path_traits_store <-
  here::here("Data/targets/traits_reference_reference/pipeline_traits_reference")

# Guard: abort rendering if the store is missing.
if (
  !fs::dir_exists(path_traits_store)
) {
  cli::cli_abort(
    c(
      "Trait pipeline store not found at: {path_traits_store}",
      "i" = stringr::str_c(
        "Run `targets::tar_make(",
        "script = 'R/02_Main_analyses/pipeline_traits_reference.R',",
        "store = 'Data/targets/traits_reference_reference/pipeline_traits_reference')` first."
      )
    )
  )
}
```

---

# Section 1 — Segment 1: Trait extraction

```{r}
#| label: s1-load

data_continental_rows <-
  targets::tar_read(
    data_continental_rows,
    store = path_traits_store
  )

vec_trait_domain_names <-
  targets::tar_read(
    vec_trait_domain_names,
    store = path_traits_store
  )

data_traits_raw <-
  targets::tar_read(
    data_traits_raw,
    store = path_traits_store
  )
```

## 1.1 Continental units

```{r}
#| label: s1-continent-table

data_continental_rows |>
  tinytable::tt()
```

## 1.2 Trait domains in VegVault

```{r}
#| label: s1-domains

tibble::tibble(trait_domain = vec_trait_domain_names) |>
  tinytable::tt()
```

## 1.3 Raw extraction overview

```{r}
#| label: s1-summary

n_taxa_raw <-
  dplyr::n_distinct(dplyr::pull(data_traits_raw, "taxon_name"))

n_domains_raw <-
  dplyr::n_distinct(dplyr::pull(data_traits_raw, "trait_domain_name"))

n_continents_raw <-
  dplyr::n_distinct(dplyr::pull(data_traits_raw, "scale_id"))

base::cat(stringr::str_glue(
  "Rows:       {base::nrow(data_traits_raw)}\n",
  "Taxa:       {n_taxa_raw}\n",
  "Domains:    {n_domains_raw}\n",
  "Continents: {n_continents_raw}\n"
))
```

## 1.4 Records per continent

```{r}
#| label: s1-records-per-continent
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 0.5

data_traits_raw |>
  dplyr::count(.data[["scale_id"]]) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = forcats::fct_reorder(.data[["scale_id"]], .data[["n"]]),
      y = .data[["n"]]
    )
  ) +
  ggplot2::geom_col(fill = "#3399cc") +
  ggplot2::coord_flip() +
  ggplot2::labs(
    x = "Continent",
    y = "N records"
  ) +
  ggplot2::theme_minimal()
```

## 1.5 Value distribution per domain (density ridges)

```{r}
#| label: s1-domain-ridges
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 1.5

data_traits_raw |>
  dplyr::filter(!base::is.na(.data[["trait_value"]])) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = .data[["trait_domain_name"]],
      y = .data[["trait_value"]],
      fill = .data[["trait_domain_name"]]
    )
  ) +
  ggplot2::geom_violin(
    alpha = 0.7,
    show.legend = FALSE
  ) +
  ggplot2::scale_y_log10() +
  ggplot2::coord_flip() +
  ggplot2::labs(
    x = NULL,
    y = "Trait value (log10 scale)"
  ) +
  ggplot2::theme_minimal()
```

---

# Section 2 — Segment 2: Raw QC flags

```{r}
#| label: s2-load

# Auto-detect most recent QC report in Data/Temp/.
vec_qc_report_paths <-
  fs::dir_ls(
    here::here("Data/Temp"),
    regexp = "trait_qc_report_\\d{4}-\\d{2}-\\d{2}\\.csv$"
  )

if (
  base::length(vec_qc_report_paths) == 0L
) {
  cli::cli_warn(
    c(
      "No trait_qc_report_*.csv found in Data/Temp/.",
      "i" = "Run the traits pipeline (Segment 2) to generate the report.",
      "i" = "Section 2 outputs will be skipped."
    )
  )

  data_qc_report <- NULL
} else {
  path_qc_report <-
    vec_qc_report_paths |>
    base::sort() |>
    utils::tail(1L)

  base::message(
    stringr::str_glue("Using QC report: {base::basename(path_qc_report)}")
  )

  data_qc_report <-
    readr::read_csv(
      path_qc_report,
      show_col_types = FALSE
    ) |>
    dplyr::mutate(
      outlier_fraction =
        .data[["n_suspected_outliers_taxon"]] / .data[["n_records"]]
    )
}
```

## 2.1 Summary per domain

```{r}
#| label: s2-summary-domain

if (
  !base::is.null(data_qc_report)
) {
  data_qc_report |>
    dplyr::group_by(.data[["trait_domain_name"]]) |>
    dplyr::summarise(
      n_groups = dplyr::n(),
      n_flagged = base::sum(
        .data[["n_suspected_outliers_taxon"]] > 0L,
        na.rm = TRUE
      ),
      mean_outlier_fraction = base::mean(
        .data[["outlier_fraction"]],
        na.rm = TRUE
      ),
      .groups = "drop"
    ) |>
    dplyr::arrange(dplyr::desc(.data[["mean_outlier_fraction"]])) |>
    tinytable::tt()
}
```

## 2.2 Outlier fraction distribution per domain

```{r}
#| label: s2-outlier-dist
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch

if (
  !base::is.null(data_qc_report)
) {
  data_qc_report |>
    dplyr::filter(!base::is.na(.data[["outlier_fraction"]])) |>
    ggplot2::ggplot(
      ggplot2::aes(
        x = forcats::fct_reorder(
          .data[["trait_domain_name"]],
          .data[["outlier_fraction"]],
          .fun = stats::median
        ),
        y = .data[["outlier_fraction"]],
        fill = .data[["trait_domain_name"]]
      )
    ) +
    ggplot2::geom_violin(
      alpha = 0.7,
      show.legend = FALSE
    ) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      x = NULL,
      y = "Outlier fraction"
    ) +
    ggplot2::theme_minimal()
}
```

## 2.3 Top-25 most flagged taxon × domain combinations

```{r}
#| label: s2-top-flagged

if (
  !base::is.null(data_qc_report)
) {
  data_qc_report |>
    dplyr::arrange(dplyr::desc(.data[["outlier_fraction"]])) |>
    utils::head(25L) |>
    dplyr::select(
      "taxon_name",
      "trait_domain_name",
      "n_records",
      "n_suspected_outliers_taxon",
      "outlier_fraction"
    ) |>
    tinytable::tt()
}
```

## 2.4 Focused domain inspector

Change `sel_domain` to inspect any specific domain.

```{r}
#| label: s2-domain-inspector

# ---- selector ----
sel_domain <- vec_trait_domain_names[1]
# ------------------

if (
  !base::is.null(data_qc_report)
) {
  data_qc_report |>
    dplyr::filter(.data[["trait_domain_name"]] == sel_domain) |>
    dplyr::arrange(dplyr::desc(.data[["outlier_fraction"]])) |>
    tinytable::tt()
}
```

---

# Section 3 — Segment 2: Corrections applied

```{r}
#| label: s3-load

data_traits_corrected <-
  targets::tar_read(
    data_traits_corrected,
    store = path_traits_store
  )
```

## 3.1 Effect of corrections

```{r}
#| label: s3-effect

n_raw <-
  base::nrow(data_traits_raw)

n_corrected <-
  base::nrow(data_traits_corrected)

base::cat(stringr::str_glue(
  "Raw records:       {n_raw}\n",
  "After corrections: {n_corrected}\n",
  "Records removed:   {n_raw - n_corrected} ",
  "({base::round((n_raw - n_corrected) / n_raw * 100, 1)}%)\n"
))
```

## 3.2 Before/after comparison for a selected taxon × domain

Change `sel_taxon` and `sel_domain_corr` to inspect any group.

```{r}
#| label: s3-before-after

# ---- selectors ----
sel_taxon <-
  data_traits_raw |>
  dplyr::slice(1L) |>
  dplyr::pull("taxon_name")
sel_domain_corr <- vec_trait_domain_names[1]
# -------------------

data_before <-
  data_traits_raw |>
  dplyr::filter(
    .data[["taxon_name"]] == sel_taxon,
    .data[["trait_domain_name"]] == sel_domain_corr
  ) |>
  dplyr::mutate(stage = "raw")

data_after <-
  data_traits_corrected |>
  dplyr::filter(
    .data[["taxon_name"]] == sel_taxon,
    .data[["trait_domain_name"]] == sel_domain_corr
  ) |>
  dplyr::mutate(stage = "corrected")

dplyr::bind_rows(data_before, data_after) |>
  dplyr::select("stage", "taxon_name", "trait_domain_name", "trait_value") |>
  dplyr::arrange(.data[["stage"]]) |>
  tinytable::tt()
```

---

# Section 4 — Segment 3: Taxonomic classification

```{r}
#| label: s4-load

data_resolution_to_finest <-
  targets::tar_read(
    data_resolution_to_finest,
    store = path_traits_store
  )

data_traits_classified <-
  targets::tar_read(
    data_traits_classified,
    store = path_traits_store
  )
```

## 4.1 Resolution overview

```{r}
#| label: s4-resolution-breakdown

n_original <-
  dplyr::n_distinct(dplyr::pull(data_resolution_to_finest, "sel_name"))

n_resolved <-
  dplyr::n_distinct(dplyr::pull(data_resolution_to_finest, "taxon_resolved"))

n_changed <-
  data_resolution_to_finest |>
  dplyr::filter(
    .data[["sel_name"]] != .data[["taxon_resolved"]]
  ) |>
  base::nrow()

base::cat(stringr::str_glue(
  "Original taxon names:  {n_original}\n",
  "Distinct resolved:     {n_resolved}\n",
  "Names changed:         {n_changed}\n",
  "Names unchanged:       {n_original - n_changed}\n"
))
```

## 4.2 Classified records overview

```{r}
#| label: s4-classified-summary

n_resolved_taxa_classified <-
  dplyr::n_distinct(dplyr::pull(data_traits_classified, "taxon_resolved"))

base::cat(stringr::str_glue(
  "Records after classification join: {base::nrow(data_traits_classified)}\n",
  "Distinct resolved taxa:            {n_resolved_taxa_classified}\n"
))
```

## 4.3 Distinct resolved taxa per continent

```{r}
#| label: s4-taxa-per-continent

data_traits_classified |>
  dplyr::group_by(.data[["scale_id"]]) |>
  dplyr::summarise(
    n_resolved_taxa = dplyr::n_distinct(.data[["taxon_resolved"]]),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data[["n_resolved_taxa"]])) |>
  tinytable::tt()
```

---

# Section 5 — Segment 4: Classified QC

```{r}
#| label: s5-load

data_traits_classified_corrected <-
  targets::tar_read(
    data_traits_classified_corrected,
    store = path_traits_store
  )
```

## 5.1 Effect of genus-level corrections

```{r}
#| label: s5-effect

n_classified <-
  base::nrow(data_traits_classified)

n_class_corrected <-
  base::nrow(data_traits_classified_corrected)

base::cat(stringr::str_glue(
  "After classification:            {n_classified}\n",
  "After genus-level corrections:   {n_class_corrected}\n",
  "Records removed:                 {n_classified - n_class_corrected}\n"
))
```

## 5.2 Top-25 most flagged genus × domain combinations

```{r}
#| label: s5-top-flagged

# Auto-detect most recent classified QC report in Data/Temp/.
vec_qc_class_paths <-
  fs::dir_ls(
    here::here("Data/Temp"),
    regexp = "trait_qc_report_classified_\\d{4}-\\d{2}-\\d{2}\\.csv$"
  )

if (
  base::length(vec_qc_class_paths) == 0L
) {
  base::message(
    "No trait_qc_report_classified_*.csv found in Data/Temp/. ",
    "Section 5.2 skipped."
  )
} else {
  path_qc_class_report <-
    vec_qc_class_paths |>
    base::sort() |>
    utils::tail(1L)

  readr::read_csv(path_qc_class_report, show_col_types = FALSE) |>
    dplyr::mutate(
      outlier_fraction =
        .data[["n_suspected_outliers_taxon"]] / .data[["n_records"]]
    ) |>
    dplyr::arrange(dplyr::desc(.data[["outlier_fraction"]])) |>
    utils::head(25L) |>
    dplyr::select(
      "taxon_name",
      "trait_domain_name",
      "n_records",
      "n_suspected_outliers_taxon",
      "outlier_fraction"
    ) |>
    tinytable::tt()
}
```

---

# Section 6 — Segment 5: Trait table assembly

```{r}
#| label: s6-load

data_traits_aggregated <-
  targets::tar_read(
    data_traits_aggregated,
    store = path_traits_store
  )

data_trait_table <-
  targets::tar_read(
    data_trait_table,
    store = path_traits_store
  )
```

## 6.1 Aggregated-values overview

```{r}
#| label: s6-aggregated-summary

n_taxa_aggregated <-
  dplyr::n_distinct(dplyr::pull(data_traits_aggregated, "taxon_resolved"))

n_domains_aggregated <-
  dplyr::n_distinct(dplyr::pull(data_traits_aggregated, "trait_domain_name"))

base::cat(stringr::str_glue(
  "Aggregated rows:  {base::nrow(data_traits_aggregated)}\n",
  "Unique taxa:      {n_taxa_aggregated}\n",
  "Unique domains:   {n_domains_aggregated}\n"
))
```

## 6.2 Trait table dimensions

```{r}
#| label: s6-table-dims

base::cat(stringr::str_glue(
  "Rows (taxa):     {base::nrow(data_trait_table)}\n",
  "Columns (traits): {base::ncol(data_trait_table) - 1L}",
  "  (excluding taxon_name)\n"
))
```

## 6.3 Per-domain coverage (non-NA fraction)

```{r}
#| label: s6-coverage
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 0.5

data_trait_table |>
  dplyr::select(-"taxon_name") |>
  purrr::map_dbl(
    .f = ~ base::mean(!base::is.na(.x))
  ) |>
  tibble::enframe(name = "trait_domain", value = "coverage") |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = forcats::fct_reorder(.data[["trait_domain"]], .data[["coverage"]]),
      y = .data[["coverage"]]
    )
  ) +
  ggplot2::geom_col(fill = "#aa5599") +
  ggplot2::coord_flip() +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::labs(
    x = NULL,
    y = "Coverage (non-NA fraction)"
  ) +
  ggplot2::theme_minimal()
```

## 6.4 Top-20 taxa with fullest trait coverage

```{r}
#| label: s6-top-taxa

data_trait_table |>
  dplyr::mutate(
    coverage = base::rowSums(!base::is.na(dplyr::pick(-"taxon_name"))) /
      (base::ncol(data_trait_table) - 1L)
  ) |>
  dplyr::select("taxon_name", "coverage") |>
  dplyr::arrange(dplyr::desc(.data[["coverage"]])) |>
  utils::head(20L) |>
  tinytable::tt()
```

## 6.5 Trait × taxon coverage heatmap (first 50 taxa)

```{r}
#| label: s6-coverage-heatmap
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch

data_trait_table |>
  utils::head(50L) |>
  tidyr::pivot_longer(
    cols = -"taxon_name",
    names_to = "trait_domain",
    values_to = "value"
  ) |>
  dplyr::mutate(
    has_value = !base::is.na(.data[["value"]])
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = .data[["trait_domain"]],
      y = .data[["taxon_name"]],
      fill = .data[["has_value"]]
    )
  ) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.3) +
  ggplot2::scale_fill_manual(
    values = base::c("TRUE" = "#3399cc", "FALSE" = "#cccccc"),
    labels = base::c("TRUE" = "Present", "FALSE" = "Missing")
  ) +
  ggplot2::labs(
    x = "Trait domain",
    y = "Taxon (first 50)",
    fill = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    axis.text.y = ggplot2::element_text(size = 5)
  )
```

---

# Section 7 — Segment 6: FT clustering

```{r}
#| label: s7-config

ft_groups_min_val <-
  get_config_value_with_fallback(
    config_section = "data_processing",
    config_key = "ft_groups_min",
    fallback_config = "project_traits_reference"
  ) |>
  base::as.integer()

ft_groups_max_val <-
  get_config_value_with_fallback(
    config_section = "data_processing",
    config_key = "ft_groups_max",
    fallback_config = "project_traits_reference"
  ) |>
  base::as.integer()

metric_val <-
  get_config_value_with_fallback(
    config_section = "data_processing",
    config_key = "ft_metric",
    fallback_config = "project_traits_reference"
  )

method_val <-
  get_config_value_with_fallback(
    config_section = "data_processing",
    config_key = "ft_method",
    fallback_config = "project_traits_reference"
  )

base::cat(stringr::str_glue(
  "ft_groups_min : {ft_groups_min_val}\n",
  "ft_groups_max : {ft_groups_max_val}\n",
  "metric        : {metric_val}\n",
  "method        : {method_val}\n"
))
```

```{r}
#| label: s7-load-continent-objects

data_continental_rows_clust <-
  targets::tar_read(
    data_continental_rows,
    store = path_traits_store
  )

vec_continent_ids <-
  dplyr::pull(data_continental_rows_clust, "scale_id")

# Branched targets (tar_map) cannot be read as a combined object via
# tar_read() when the branches return non-stackable types (dist, hclust).
# Read each branch individually using branch names from the pipeline metadata,
# then collect into a named list keyed by continent.
data_meta_pipeline <-
  targets::tar_meta(store = path_traits_store)

# Read all branches of a pattern target as a named list (one element per
# continent, in the same order as vec_continent_ids).
continent_branches <- function(pattern_name) {
  vec_children <-
    data_meta_pipeline |>
    dplyr::filter(.data[["name"]] == pattern_name) |>
    dplyr::pull("children") |>
    purrr::chuck(1L)

  vec_children |>
    rlang::set_names(vec_continent_ids) |>
    purrr::map(
      .f = ~ targets::tar_read_raw(.x, store = path_traits_store)
    )
}

list_continent_traits <-
  continent_branches("data_continent_traits")

list_dist <-
  continent_branches("dist_continent")

list_hclust <-
  continent_branches("hclust_continent")

list_ft_groups_chosen <-
  continent_branches("ft_groups_chosen_continent")

list_ft_result <-
  continent_branches("ft_result_continent")
```

## 7.1 Trait matrix dimensions per continent

```{r}
#| label: s7-matrix-dims

list_continent_traits |>
  purrr::imap(
    .f = ~ tibble::tibble(
      continent = .y,
      n_taxa = base::nrow(.x),
      n_traits = base::ncol(.x) - 1L
    )
  ) |>
  purrr::list_rbind() |>
  tinytable::tt()
```

## 7.2 Trait × taxon NA coverage per continent

```{r}
#| label: s7-na-heatmap
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 1.2

list_continent_traits |>
  purrr::imap(
    .f = ~ {
      .x |>
        utils::head(60L) |>
        tidyr::pivot_longer(
          cols = -dplyr::any_of("taxon_name"),
          names_to = "trait_domain",
          values_to = "value"
        ) |>
        dplyr::mutate(
          has_value = !base::is.na(.data[["value"]])
        ) |>
        ggplot2::ggplot(
          ggplot2::aes(
            x = .data[["trait_domain"]],
            y = .data[["taxon_name"]],
            fill = .data[["has_value"]]
          )
        ) +
        ggplot2::geom_tile(colour = "white", linewidth = 0.25) +
        ggplot2::scale_fill_manual(
          values = base::c("TRUE" = "#3399cc", "FALSE" = "#cccccc"),
          labels = base::c("TRUE" = "Present", "FALSE" = "Missing")
        ) +
        ggplot2::labs(
          title = stringr::str_glue("Coverage: {.y} (first 60 taxa)"),
          x = "Trait domain",
          y = NULL,
          fill = NULL
        ) +
        ggplot2::theme_minimal() +
        ggplot2::theme(
          axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
          axis.text.y = ggplot2::element_text(size = 5)
        )
    }
  )
```

## 7.3 Silhouette width vs. k (optimal k highlighted)

Sweeps the full k range using the stored `dist` and `hclust` objects.

```{r}
#| label: s7-silhouette-curve
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch

base::set.seed(900723)

vec_continent_ids |>
  purrr::map(
    .f = ~ {
      continent <-
        .x

      dist_mat <-
        purrr::chuck(list_dist, continent)

      list_hclust_obj <-
        purrr::chuck(list_hclust, continent)

      n_obs <-
        purrr::chuck(list_hclust_obj, "order") |>
        base::length()

      k_max_clamped <-
        base::min(ft_groups_max_val, n_obs - 1L)

      k_min_clamped <-
        base::min(ft_groups_min_val, k_max_clamped)

      vec_k <-
        base::seq(k_min_clamped, k_max_clamped)

      vec_sil_mean <-
        vec_k |>
        purrr::map_dbl(
          .f = ~ {
            vec_cut <-
              stats::cutree(list_hclust_obj, k = .x)

            sil <-
              cluster::silhouette(vec_cut, dist_mat)

            base::mean(sil[, "sil_width"])
          }
        )

      k_optimal <-
        purrr::chuck(list_ft_groups_chosen, continent)

      tibble::tibble(k = vec_k, mean_silhouette_width = vec_sil_mean) |>
        ggplot2::ggplot(
          ggplot2::aes(
            x = .data[["k"]],
            y = .data[["mean_silhouette_width"]]
          )
        ) +
        ggplot2::geom_line(colour = "#555555") +
        ggplot2::geom_point(size = 2, colour = "#555555") +
        ggplot2::geom_vline(
          xintercept = k_optimal,
          linetype = "dashed",
          colour = "#cc3333"
        ) +
        ggplot2::annotate(
          "text",
          x = k_optimal + 0.3,
          y = base::max(vec_sil_mean),
          label = stringr::str_glue("k = {k_optimal}"),
          hjust = 0,
          colour = "#cc3333"
        ) +
        ggplot2::labs(
          title = stringr::str_glue("Silhouette curve: {continent}"),
          x = "Number of FT groups (k)",
          y = "Mean silhouette width"
        ) +
        ggplot2::theme_minimal()
    }
  )
```

## 7.4 FT membership summary per continent

```{r}
#| label: s7-ft-membership

list_ft_result |>
  purrr::imap(
    .f = ~ .x |>
      dplyr::group_by(.data[["functional_type"]]) |>
      dplyr::summarise(
        n_taxa = dplyr::n(),
        mean_silhouette_width = base::mean(
          .data[["silhouette_width"]],
          na.rm = TRUE
        ),
        median_silhouette_width = stats::median(
          .data[["silhouette_width"]],
          na.rm = TRUE
        ),
        top_3_taxa = stringr::str_c(
          utils::head(.data[["taxon_name"]], 3L),
          collapse = "; "
        ),
        .groups = "drop"
      ) |>
      dplyr::mutate(continent = .y) |>
      dplyr::relocate("continent")
  ) |>
  purrr::list_rbind() |>
  dplyr::arrange(
    .data[["continent"]],
    .data[["functional_type"]]
  ) |>
  tinytable::tt()
```

## 7.5 Silhouette width distribution per FT

```{r}
#| label: s7-sil-boxplot
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch

list_ft_result |>
  purrr::imap(
    .f = ~ dplyr::mutate(.x, continent = .y)
  ) |>
  purrr::list_rbind() |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = base::factor(.data[["functional_type"]]),
      y = .data[["silhouette_width"]],
      fill = .data[["continent"]]
    )
  ) +
  ggplot2::geom_boxplot(
    alpha = 0.6,
    outlier.size = 0.8
  ) +
  ggplot2::geom_hline(
    yintercept = 0,
    linetype = "dashed",
    colour = "#cc3333"
  ) +
  ggplot2::facet_wrap(
    ggplot2::vars(.data[["continent"]]),
    scales = "free_x"
  ) +
  ggplot2::labs(
    x = "Functional type",
    y = "Silhouette width",
    fill = "Continent"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "none"
  )
```

## 7.6 Per-FT trait profile (median trait value per FT)

```{r}
#| label: s7-ft-trait-profile
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 1.2

# ---- selector ----
sel_continent_profile <- vec_continent_ids[1]
# ------------------

data_trait_mat <-
  purrr::chuck(list_continent_traits, sel_continent_profile)

data_ft_assignments <-
  purrr::chuck(list_ft_result, sel_continent_profile)

data_trait_mat |>
  dplyr::inner_join(
    data_ft_assignments |>
      dplyr::select("taxon_name", "functional_type"),
    by = dplyr::join_by(taxon_name)
  ) |>
  dplyr::select(-"taxon_name") |>
  dplyr::group_by(.data[["functional_type"]]) |>
  dplyr::summarise(
    dplyr::across(
      dplyr::where(base::is.numeric),
      .fns = ~ stats::median(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols = -"functional_type",
    names_to = "trait_domain",
    values_to = "median_value"
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = .data[["trait_domain"]],
      y = base::factor(.data[["functional_type"]]),
      fill = .data[["median_value"]]
    )
  ) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.3) +
  ggplot2::scale_fill_viridis_c(
    option = "plasma",
    na.value = "#dddddd"
  ) +
  ggplot2::labs(
    title = stringr::str_glue(
      "Per-FT trait profile — {sel_continent_profile}"
    ),
    x = "Trait domain",
    y = "Functional type",
    fill = "Median value"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

## 7.7 Dendrogram coloured by FT assignment

```{r}
#| label: s7-dendrogram
#| fig-width: !expr fig_width_inch
#| fig-height: !expr fig_height_inch * 1.5

# ---- selector ----
sel_continent_dendro <- vec_continent_ids[1]
# ------------------

list_hclust_dendro <-
  purrr::chuck(list_hclust, sel_continent_dendro)

data_ft_assignments_dendro <-
  purrr::chuck(list_ft_result, sel_continent_dendro) |>
  dplyr::select("taxon_name", "functional_type")

k_optimal_dendro <-
  purrr::chuck(list_ft_groups_chosen, sel_continent_dendro)

# Build colour palette (one per FT).
vec_ft_labels <-
  data_ft_assignments_dendro |>
  dplyr::pull("functional_type") |>
  base::unique() |>
  base::sort()

vec_palette <-
  grDevices::hcl.colors(
    n = base::length(vec_ft_labels),
    palette = "Dark 3"
  ) |>
  rlang::set_names(vec_ft_labels)

# Map FT colour to leaf order.
vec_hclust_order <-
  purrr::chuck(list_hclust_dendro, "order")

vec_hclust_labels <-
  purrr::chuck(list_hclust_dendro, "labels")

vec_leaf_order <-
  vec_hclust_labels[vec_hclust_order]

vec_leaf_colours <-
  data_ft_assignments_dendro |>
  dplyr::filter(.data[["taxon_name"]] %in% vec_leaf_order) |>
  dplyr::mutate(
    colour = purrr::map_chr(
      .data[["functional_type"]],
      .f = ~ purrr::chuck(vec_palette, base::as.character(.x))
    )
  ) |>
  dplyr::arrange(base::match(.data[["taxon_name"]], vec_leaf_order)) |>
  dplyr::pull("colour")

graphics::plot(
  stats::as.dendrogram(list_hclust_dendro),
  horiz = TRUE,
  leaflab = "none",
  main = stringr::str_glue(
    "Dendrogram — {sel_continent_dendro} (k = {k_optimal_dendro})"
  )
)
```

---

*Notebook generated on `r Sys.Date()` from the trait pipeline store at `Data/targets/traits_reference_reference/pipeline_traits_reference`.*
