---
title: "MOTHER Ecological Biosurveillance"
subtitle: "IAVS 2026 proof of concept"
author: "BIODYNAMICS vegetation co-occurrence"
format:
  revealjs:
    theme:
      - default
      - mother.scss
    slide-width: 1600
    slide-height: 900
    slide-number: true
    embed-resources: true
    center-title-slide: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
here::i_am("Documentation/Presentations/IAVS_2026/index.qmd")
source(
  here::here("R/Supplementary/Presentation/theme_mother.R")
)
```

# MOTHER ECOLOGICAL BIOSURVEILLANCE {.mother-title}

::: {.terminal-status}
SYSTEM: **ONLINE** / SESSION: IAVS_2026 / MODE: VEGETATION CO-OCCURRENCE
:::

::: {.mother-says .terminal-cursor}
Awaiting research query. Climate archive and fossil pollen matrices are mounted.
:::

::: {.terminal-grid}
::: {.terminal-panel}
[Archive]{.label}

[VegVault]{.value}
:::

::: {.terminal-panel}
[Signal]{.label}
[Co-occurrence]{.value}
:::

::: {.terminal-panel}
[Slot]{.label}

[12 min]{.value}
:::
:::

## Can MOTHER separate climate response from association? {.query-slide}

::: {.mother-says}
Query accepted. Compare the observed co-occurrence structure with abiotic predictors, spatial structure, and time-slice dynamics.
:::

| Input stream | Diagnostic role | Status |
|---|---|---|
| Fossil pollen archives | Plant community signal | mounted |
| Palaeoclimate rasters | Shared abiotic response | mounted |
| Spatial and temporal filters | Scale sensitivity | pending |

## Prototype visual output {.result-slide}

```{r}
#| label: prototype-figure
#| fig-width: 9
#| fig-height: 4.6
#| out-width: "82%"

prototype_data <- data.frame(
  age_kyr = rep(c(20, 15, 10, 5, 0), 3),
  signal = c(
    0.18, 0.24, 0.31, 0.36, 0.39,
    0.12, 0.20, 0.27, 0.29, 0.33,
    0.08, 0.14, 0.22, 0.25, 0.30
  ),
  component = rep(
    c("Abiotic response", "Spatial structure", "Residual association"),
    each = 5
  )
)

ggplot2::ggplot(
  prototype_data,
  ggplot2::aes(x = age_kyr, y = signal, colour = component)
) +
  ggplot2::geom_line(linewidth = 1.15) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_x_reverse(breaks = c(20, 15, 10, 5, 0)) +
  scale_colour_mother_discrete() +
  ggplot2::labs(
    title = "Prototype variance signal",
    subtitle = "Rendered with theme_mother() and design_config.json colors",
    x = "Age before present (kyr)",
    y = "Explained signal",
    colour = "Component"
  ) +
  theme_mother(base_size = 18)
```

::: {.terminal-status}
CONFIG SOURCE: **design_config.json** / THEME SOURCE: **theme_mother()**
:::
