---
date: 2025/05/15
date-format: long
date-modified: last-modified
listing: 
  - id: functions
    max-description-length: 100
    fields: [title]
    contents: "Functions"
    type: grid
    grid-item-border: true
---

# BIODYNAMICS - Vegetation Co-occurrence

This is a project to study the co-occurrence of vegetation across space and time.

The repo is part of the [BIODYNAMICS](https://bit.ly/BIODYNAMICS) project.

See [detailed description of the project](/website/about.qmd) for more information.

The latest information are always available on the [project website](/index.qmd).

# Functions

This project is coded in [R](https://www.r-project.org/) and uses {[targets](https://docs.ropensci.org/targets/)} pipeline to run the analysis. 

## Code coverage

```{r}
#| label: estimate_test_coverage
#| echo: false
#| output: false
#| message: false
#| warning: false
#| error: false

library(here)

here::i_am("website/Documentation/documentation.qmd")

source(
  here::here("R/___setup_project___.R")
)

source(
  here::here("R/03_Supplementary_analyses/Estimate_test_coverage.R")
)
```

```{r}
#| label: make_test_coverage_quarto
#| echo: false
#| output: false
#| message: false
#| warning: false
#| error: false

base_name <-
  fs::path_file(
    here::here(
      "website/Documentation/code_coverage_report.qmd"
    )
  )

name_no_ext <-
  fs::path_ext_remove(base_name)

# Read the content of the HTML file
html_content <-
  readLines(
    here::here(
      "docs/Documentation/Functions_test_coverage/covr_report.html"
    ),
    warn = FALSE
  )

# Define the content of the .qmd file
qmd_content <-
  c(
    "---",
    "format: html",
    "date: 2025/05/15",
    "date-format: long",
    "date-modified: last-modified",
    "title: Code coverage report",
    "---",
    "",
    html_content
  )

qmd_content[9] <-
  qmd_content[9] %>%
  stringr::str_remove(
    pattern = "^<!DOCTYPE html>"
  ) 

# Write the content to a .qmd file
writeLines(
  qmd_content,
  here::here(
    "website/Documentation/code_coverage_report.qmd"
  )
)
```


In order to check issues during the development of the code, we are writing test using the {[testthat](https://testthat.r-lib.org/)} package. Then we are using the {[covr](https://covr.r-lib.org/)} package to check the code coverage. The code coverage is calculated for each function and the overall code coverage is calculated as well.

Current code coverage is: ![](https://img.shields.io/badge/dynamic/json?url=https://raw.githubusercontent.com/OndrejMottl/BIODYNAMICS-vegetation-cooccurrence/main/Documentation/Functions_test_coverage/covr_report_summary.json&query=$.value&label=codecov&color=orange&style=flat-square&suffix=%25)


The full report of the {[covr](https://covr.r-lib.org/)} package can be found in on the [code coverage report page](/website/Documentation/code_coverage_report.qmd)

## Documentation of individual functions

Each function used in the project is documented using the [roxygen2](https://roxygen2.r-lib.org/) package. Here we present the list of functions used in the project:

::: {#functions}
:::