---
title: "Trait review automation pilot"
subtitle: "Diaspore mass and stem specific density"
date: today
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(readr)
library(knitr)

path_pilot <-
  here::here("Data/Temp/Trait_corrections/raw/automation")
data_recommendations <-
  read_csv(
    file.path(path_pilot, "pilot_candidate_recommendations.csv"),
    show_col_types = FALSE
  )
data_comparison <-
  read_csv(
    file.path(path_pilot, "pilot_historical_comparison.csv"),
    show_col_types = FALSE
  )
data_proposals <-
  read_csv(
    file.path(path_pilot, "pilot_proposal_diagnostics.csv"),
    show_col_types = FALSE
  )
data_records <-
  read_csv(
    file.path(path_pilot, "pilot_record_evidence.csv"),
    show_col_types = FALSE
  )

n_candidates <- nrow(data_recommendations)
n_policy <- sum(
  data_recommendations$acceptance_eligibility ==
    "eligible_policy_acceptance"
)
n_agent <- sum(
  data_recommendations$acceptance_eligibility ==
    "requires_agent_review"
)
n_batches <- n_distinct(
  data_recommendations$agent_batch_id,
  na.rm = TRUE
)
```

## Outcome

This was a non-mutating dry run: it did not edit either canonical decision file and did not apply any correction to trait records.

The pilot evaluated `r format(n_candidates, big.mark = ",")` current taxon-trait candidates against `r format(nrow(data_records), big.mark = ",")` matching source records.
It found `r n_policy` candidates eligible for the approved policy-based no-action route because their current record count, median, and IQR remain sufficiently stable relative to the archived review.
The remaining `r n_agent` candidates are organized into `r n_batches` batches of at most 25 for agent evidence review.

```{r}
data_recommendations |>
  count(
    trait_domain_name,
    acceptance_eligibility,
    name = "candidates"
  ) |>
  mutate(
    route = recode(
      acceptance_eligibility,
      eligible_policy_acceptance = "Policy no-action",
      requires_agent_review = "Agent evidence review"
    )
  ) |>
  select(`Trait domain` = trait_domain_name, Route = route, Candidates = candidates) |>
  kable()
```

## Why candidates were routed this way

The deterministic layer is intentionally conservative.
It can recommend no action when archived no-action evidence remains stable, and it can identify invalid non-positive values in domains that must be positive.
It does not automatically accept recovered selectors, inferred unit patterns, or historically reviewed groups whose current distributions have drifted.

```{r}
data_recommendations |>
  count(
    trait_domain_name,
    deterministic_outcome,
    name = "candidates"
  ) |>
  arrange(trait_domain_name, desc(candidates)) |>
  mutate(
    deterministic_outcome = recode(
      deterministic_outcome,
      retain_historical_no_action = "Stable historical no-action",
      investigate_historical_drift = "Historical summary drift",
      investigate_unit_pattern = "Possible unit pattern",
      validate_recovered_proposal = "Recovered selector to validate",
      retain_flagged_uncertain = "Plausible but uncertain; retain flagged",
      exclude_invalid_values = "Invalid non-positive value"
    )
  ) |>
  select(
    `Trait domain` = trait_domain_name,
    `Pilot outcome` = deterministic_outcome,
    Candidates = candidates
  ) |>
  kable()
```

All `r nrow(data_proposals)` recovered selector drafts in these two domains still match at least one current record.
They remain proposals because matching a selector does not establish that its correction is scientifically correct.

## Comparison with the archived review

The archived report is not treated as an infallible answer key.
The comparison checks whether the new recommendation agrees with the broad historical heuristic and whether the underlying record count and rounded distribution summary remain stable.
`changed_or_more_conservative` therefore includes both genuine data drift and cases where the new policy requires stronger evidence than the old heuristic.

```{r}
data_comparison |>
  count(
    trait_domain_name,
    historical_comparison,
    name = "candidates"
  ) |>
  arrange(trait_domain_name, historical_comparison) |>
  select(
    `Trait domain` = trait_domain_name,
    Comparison = historical_comparison,
    Candidates = candidates
  ) |>
  kable()
```

Five current candidates have no exact candidate identifier in the archived reports.
This reflects changed taxon-domain membership rather than five wholly new domains: the archived and current sets must be compared in both directions before interpreting additions and removals.

## Agent batches

Only candidates that still require evidence review receive an `agent_batch_id`.
Policy-eligible candidates are deliberately excluded from the agent workload.

```{r}
data_recommendations |>
  filter(!is.na(agent_batch_id)) |>
  count(trait_domain_name, agent_batch_id, name = "candidates") |>
  select(
    `Trait domain` = trait_domain_name,
    `Agent batch` = agent_batch_id,
    Candidates = candidates
  ) |>
  kable()
```

The highest-impact unresolved candidates should be reviewed first.

```{r}
data_recommendations |>
  filter(acceptance_eligibility == "requires_agent_review") |>
  arrange(desc(candidate_impact_score)) |>
  slice_head(n = 20L) |>
  transmute(
    `Trait domain` = trait_domain_name,
    Taxon = taxon_name,
    `Agent batch` = agent_batch_id,
    Outcome = deterministic_outcome,
    Impact = round(candidate_impact_score, 3)
  ) |>
  kable()
```

## Recommended next step

Run `.ai/agents/trait-correction-pilot-reviewer.agent.md` twice independently for each of the three batches, using roles `reviewer_a` and `reviewer_b`, and then run it once as `adjudicator`.
Begin with `diaspore_mass_01`; it is the smallest complete domain batch and provides a clean calibration before the two stem-density batches.
Do not copy any agent output into the canonical decisions automatically.
After the adjudicated pilot is inspected, decide whether to adopt a formal policy-authority status in the decision contract and extend the workflow to the other four domains.
