---
title: "Representative review of programmatic scale proposals"
subtitle: "Human decision aid for the conservative policy batch"
format:
  gfm:
    toc: true
    wrap: none
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
here::i_am(
  stringr::str_c(
    "R/03_Supplementary_analyses/One_time/Trait_corrections/",
    "trait_review_programmatic_spot_check_report.qmd"
  )
)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
base::options(knitr.kable.NA = "")

path_triage <-
  here::here(
    "Data/Temp/Trait_corrections/raw/programmatic_triage"
  )
data_spot_check_raw <-
  readr::read_csv(
    base::file.path(
      path_triage,
      "trait_review_programmatic_spot_check.csv"
    ),
    show_col_types = FALSE
  )
data_spot_check_decisions <-
  readr::read_csv(
    here::here(
      "Data/Input/Trait_corrections/",
      "trait_review_programmatic_spot_check_decisions.csv"
    ),
    show_col_types = FALSE
  )
data_spot_check <-
  data_spot_check_raw |>
  dplyr::select(
    -"reviewer_decision",
    -"reviewer_notes"
  ) |>
  dplyr::inner_join(
    data_spot_check_decisions |>
      dplyr::select(
        "spot_check_id",
        reviewer_decision = "review_outcome",
        reviewer_notes = "reviewer_note"
      ),
    by = dplyr::join_by(spot_check_id),
    unmatched = "error",
    relationship = "one-to-one"
  )
data_spot_check_records <-
  readr::read_csv(
    base::file.path(
      path_triage,
      "trait_review_programmatic_spot_check_records.csv"
    ),
    show_col_types = FALSE
  )
data_all_proposals <-
  readr::read_csv(
    base::file.path(
      path_triage,
      "trait_review_programmatic_decision_proposals.csv"
    ),
    show_col_types = FALSE
  )
data_stratum_assessment <-
  readr::read_csv(
    base::file.path(
      path_triage,
      "trait_review_programmatic_spot_check_stratum_assessment.csv"
    ),
    show_col_types = FALSE
  )
data_scale_approval_audit <-
  readr::read_csv(
    here::here(
      "Outputs/Reports/Trait_corrections/raw/",
      "trait_review_supported_scale_approval_audit.csv"
    ),
    show_col_types = FALSE
  )

n_spot_check <-
  base::nrow(data_spot_check)
n_all_scale_proposals <-
  base::sum(data_all_proposals[["action"]] == "scale")
n_spot_check_records <-
  base::sum(data_spot_check_records[["selected_for_scaling"]])
n_source_cases <-
  base::sum(
    data_spot_check[["reconciliation_outcome"]] ==
      "propose_scale_corroborated_source"
  )
n_threshold_cases <-
  n_spot_check - n_source_cases
n_approved_cases <-
  base::sum(data_spot_check[["reviewer_decision"]] == "approve")
n_rejected_cases <-
  base::sum(data_spot_check[["reviewer_decision"]] == "reject")
n_targeted_cases <-
  base::sum(
    data_spot_check[["reviewer_decision"]] ==
      "needs_targeted_review"
  )
n_supported_proposals <-
  base::sum(data_stratum_assessment[["n_supported_proposals"]])
n_supported_records <-
  data_stratum_assessment |>
  dplyr::filter(
    .data[["expansion_assessment"]] ==
      "supports_conservative_expansion"
  ) |>
  dplyr::summarise(
    n_records = base::sum(.data[["n_records"]])
  ) |>
  dplyr::pull(.data[["n_records"]])
n_recorded_scale_approvals <-
  base::nrow(data_scale_approval_audit)
n_recorded_scaled_records <-
  base::sum(data_scale_approval_audit[["n_scaled"]])
n_individual_approvals_outside_expansion <-
  data_spot_check |>
  dplyr::filter(
    .data[["reviewer_decision"]] == "approve",
    !.data[["candidate_id"]] %in%
      data_scale_approval_audit[["candidate_id"]]
  ) |>
  base::nrow()

format_number <- function(value) {
  if (
    base::is.na(value)
  ) {
    return("not set")
  }
  return(
    base::format(
      value,
      digits = 5L,
      scientific = FALSE,
      trim = TRUE
    )
  )
}

format_selector <- function(data_case) {
  trait_name <- data_case[["trait_name"]][[1L]]
  dataset_id <- data_case[["dataset_id"]][[1L]]
  value_lower <- data_case[["value_lower"]][[1L]]
  value_upper <- data_case[["value_upper"]][[1L]]
  vec_parts <- character()
  if (
    !base::is.na(trait_name)
  ) {
    vec_parts <-
      base::c(
        vec_parts,
        stringr::str_glue("trait name = {trait_name}")
      )
  }
  if (
    !base::is.na(dataset_id)
  ) {
    vec_parts <-
      base::c(
        vec_parts,
        stringr::str_glue("dataset ID = {dataset_id}")
      )
  }
  if (
    !base::is.na(value_lower)
  ) {
    comparator <-
      dplyr::if_else(
        data_case[["value_lower_inclusive"]][[1L]],
        ">=",
        ">"
      )
    vec_parts <-
      base::c(
        vec_parts,
        stringr::str_glue(
          "value {comparator} {format_number(value_lower)}"
        )
      )
  }
  if (
    !base::is.na(value_upper)
  ) {
    comparator <-
      dplyr::if_else(
        data_case[["value_upper_inclusive"]][[1L]],
        "<=",
        "<"
      )
    vec_parts <-
      base::c(
        vec_parts,
        stringr::str_glue(
          "value {comparator} {format_number(value_upper)}"
        )
      )
  }
  if (
    base::length(vec_parts) == 0L
  ) {
    return("all records in the taxon-domain candidate")
  }
  return(stringr::str_c(vec_parts, collapse = "; "))
}
```

## Purpose and decision boundary

This report presents `r n_spot_check` deliberately varied cases from the `r n_all_scale_proposals` unapproved programmatic scale proposals.
It contains `r n_threshold_cases` threshold-based cases and both `r n_source_cases` independently corroborated dataset-specific cases.
Together, the displayed rules would scale `r scales::comma(n_spot_check_records)` current records if approved.

The sample is a policy check, not a random estimate of an error rate.
It covers every factor-and-direction stratum, all three trait domains represented among the proposals, small and large record impacts, contrasting before/after gaps, and stable-hash fill cases.
No proposal shown here has been approved, added to the canonical decision file, or applied to production data.

For each case, decide whether the selector isolates a plausible unit mismatch and whether multiplying only those selected records by the proposed factor makes the selected and unselected distributions scientifically coherent.
Use `approve`, `reject`, or `needs targeted review` as the reviewer decision.

## Coverage

```{r coverage}
data_coverage <-
  data_spot_check |>
  dplyr::count(
    .data[["trait_domain_name"]],
    .data[["reconciliation_outcome"]],
    .data[["scale_factor"]],
    .data[["threshold_direction"]],
    name = "Cases"
  ) |>
  dplyr::rename(
    `Trait domain` = "trait_domain_name",
    `Evidence rule` = "reconciliation_outcome",
    `Scale factor` = "scale_factor",
    `Threshold direction` = "threshold_direction"
  )

knitr::kable(data_coverage)
```

## Review roster

```{r roster}
data_roster <-
  data_spot_check |>
  dplyr::select(
    `Case` = "spot_check_id",
    `Taxon` = "taxon_name",
    `Trait domain` = "trait_domain_name",
    `Factor` = "scale_factor",
    `Records changed` = "proposed_record_count",
    `Selection reason` = "selection_reason",
    `Reviewer decision` = "reviewer_decision",
    `Reviewer notes` = "reviewer_notes"
  )

knitr::kable(data_roster)
```

## Recorded decisions

The durable review record contains `r n_approved_cases` approvals, `r n_rejected_cases` rejection, and `r n_targeted_cases` cases requiring targeted investigation.
These outcomes are stored separately from the canonical correction rules and do not themselves apply any correction.

## Conservative expansion assessment

A rule stratum supports conservative expansion only when every representative reviewed from that stratum was approved.
Mixed strata, rejected strata, and the dataset-specific proposals are not expanded.
Under this rule, the reviewed sample supported expansion to `r n_supported_proposals` of the `r n_all_scale_proposals` proposals, affecting `r scales::comma(n_supported_records)` records.
The project owner subsequently approved this expansion.
Exactly `r n_recorded_scale_approvals` candidate-level scale rules are now recorded in the canonical raw decision file, and their application audit confirms `r scales::comma(n_recorded_scaled_records)` scaled records.

```{r assessment}
data_assessment_display <-
  data_stratum_assessment |>
  dplyr::select(
    `Trait domain` = "trait_domain_name",
    `Rule kind` = "rule_kind",
    `Direction` = "threshold_direction",
    `Factor` = "scale_factor",
    `All proposals` = "n_proposals",
    `Reviewed` = "n_reviewed",
    `Approved` = "n_approved",
    `Rejected` = "n_rejected",
    `Targeted review` = "n_targeted",
    `Expansion assessment` = "expansion_assessment"
  )

knitr::kable(data_assessment_display)
```

## Individual cases

```{r cases, results='asis'}
for (
  index_case in base::seq_len(base::nrow(data_spot_check))
) {
  data_case <-
    data_spot_check[index_case, , drop = FALSE]
  candidate_id <- data_case[["candidate_id"]][[1L]]
  data_case_records <-
    data_spot_check_records |>
    dplyr::filter(
      .data[["candidate_id"]] == .env[["candidate_id"]]
    )
  data_case_summary <-
    data_case_records |>
    dplyr::group_by(.data[["selected_for_scaling"]]) |>
    dplyr::summarise(
      Records = dplyr::n(),
      `Before minimum` = base::min(.data[["trait_value"]]),
      `Before median` = stats::median(.data[["trait_value"]]),
      `Before maximum` = base::max(.data[["trait_value"]]),
      `After minimum` = base::min(.data[["trait_value_after"]]),
      `After median` = stats::median(.data[["trait_value_after"]]),
      `After maximum` = base::max(.data[["trait_value_after"]]),
      .groups = "drop"
    ) |>
    dplyr::rename(
      `Scaled` = "selected_for_scaling"
    )
  data_case_sources <-
    data_case_records |>
    dplyr::group_by(
      .data[["dataset_id"]],
      .data[["dataset_name"]],
      .data[["trait_name"]],
      .data[["selected_for_scaling"]]
    ) |>
    dplyr::summarise(
      Records = dplyr::n(),
      Median = stats::median(.data[["trait_value"]]),
      `Median after` = stats::median(.data[["trait_value_after"]]),
      .groups = "drop"
    ) |>
    dplyr::arrange(
      dplyr::desc(.data[["selected_for_scaling"]]),
      dplyr::desc(.data[["Records"]]),
      .data[["dataset_id"]]
    ) |>
    dplyr::slice_head(n = 5L) |>
    dplyr::rename(
      `Dataset ID` = "dataset_id",
      `Dataset` = "dataset_name",
      `Trait name` = "trait_name",
      `Scaled` = "selected_for_scaling"
    )

  base::cat(
    stringr::str_glue(
      "\n### {data_case[['spot_check_id']][[1L]]}: ",
      "{data_case[['taxon_name']][[1L]]} - ",
      "{data_case[['trait_domain_name']][[1L]]}\n\n"
    )
  )
  base::cat(
    stringr::str_glue(
      "**Why selected:** ",
      "{data_case[['selection_reason']][[1L]]}.\n\n"
    )
  )
  base::cat(
    stringr::str_glue(
      "**Proposed rule:** multiply by ",
      "{format_number(data_case[['scale_factor']][[1L]])}; ",
      "selector: {format_selector(data_case)}.\n\n"
    )
  )
  base::cat(
    stringr::str_glue(
      "**Submitted instruction:** ",
      "{data_case[['notes_lower']][[1L]]}; archived source row ",
      "{data_case[['source_rows']][[1L]]}.\n\n"
    )
  )
  if (
    data_case[["reconciliation_outcome"]][[1L]] ==
      "propose_scale_validated_threshold"
  ) {
    base::cat(
      stringr::str_glue(
        "**Quantitative check:** selected/unselected records = ",
        "{data_case[['n_selected']][[1L]]}/",
        "{data_case[['n_unselected']][[1L]]}; median log10 gap ",
        "before = ",
        "{format_number(data_case[['threshold_gap_before']][[1L]])}, ",
        "after = ",
        "{format_number(data_case[['threshold_gap_after']][[1L]])}.\n\n"
      )
    )
  } else {
    base::cat(
      stringr::str_glue(
        "**Quantitative check:** low/high source records = ",
        "{data_case[['n_low_records']][[1L]]}/",
        "{data_case[['n_high_records']][[1L]]}; the submitted factor ",
        "matches the current source-median ratio within 5%.\n\n"
      )
    )
  }
  reviewer_decision_label <-
    stringr::str_replace_all(
      data_case[["reviewer_decision"]][[1L]],
      "_",
      " "
    )
  reviewer_notes <-
    dplyr::if_else(
      base::is.na(data_case[["reviewer_notes"]][[1L]]) |
        !base::nzchar(data_case[["reviewer_notes"]][[1L]]),
      "None recorded.",
      data_case[["reviewer_notes"]][[1L]]
    )
  base::cat(
    stringr::str_glue(
      "**Reviewer decision:** {reviewer_decision_label}\n\n"
    )
  )
  base::cat(
    stringr::str_glue(
      "**Reviewer notes:** {reviewer_notes}\n\n"
    )
  )
  base::cat("Distribution summary by selector status:\n\n")
  base::print(
    knitr::kable(
      data_case_summary,
      digits = 5L
    )
  )
  base::cat("\nFive largest or selected source groups:\n\n")
  base::print(
    knitr::kable(
      data_case_sources,
      digits = 5L
    )
  )
  base::cat("\n")
}
```

## How to use the result

The conservative 50-rule expansion is complete.
There are `r n_individual_approvals_outside_expansion` individually approved representatives inside mixed strata that were not part of the batch authorization and therefore remain outside the canonical correction file.
The next decision is whether to promote those cases individually before investigating the six targeted-review cases.
