---
title: "Trait review policy proposal"
subtitle: "All-domain raw-stage triage"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

path_policy <-
  here::here("Data/Temp/Trait_corrections/raw/policy")

data_candidates <-
  readr::read_csv(
    base::file.path(path_policy, "trait_review_policy_candidates.csv"),
    show_col_types = FALSE
  )
data_recommendations <-
  readr::read_csv(
    base::file.path(
      path_policy,
      "trait_review_policy_candidate_recommendations.csv"
    ),
    show_col_types = FALSE
  )
data_proposals <-
  readr::read_csv(
    base::file.path(path_policy, "trait_review_policy_proposals.csv"),
    show_col_types = FALSE
  )
data_agent_queue <-
  readr::read_csv(
    base::file.path(path_policy, "trait_review_policy_agent_queue.csv"),
    show_col_types = FALSE
  )
data_distribution_scenarios <-
  readr::read_csv(
    base::file.path(
      path_policy,
      "trait_review_policy_distribution_scenarios.csv"
    ),
    show_col_types = FALSE
  )

n_candidates <- base::nrow(data_candidates)
n_proposals <- base::nrow(data_proposals)
n_none_proposals <-
  base::sum(data_proposals[["action"]] == "none")
n_exclusion_proposals <-
  base::sum(data_proposals[["action"]] == "exclude")
n_investigation <-
  base::sum(
    data_recommendations[["acceptance_eligibility"]] ==
      "requires_agent_review"
  )
n_agent_queue <- base::nrow(data_agent_queue)
n_investigation_deferred <- n_investigation - n_agent_queue
n_records_proposed_excluded <-
  base::sum(
    data_distribution_scenarios[["n_records_proposed_excluded"]]
  )
```

## Decision requested

This report proposes a conservative policy for completing raw-stage trait review without treating statistical outliers as measurement errors.

No row in this report is approved, and neither canonical decision file has been edited.

The proposed policy would retain a flagged taxon-domain group when no objective error evidence or unresolved submitted concern exists, exclude only exact zero values in strictly positive trait domains, and send negative, non-finite, source-unit, recovered-selector, and unresolved submitted concerns to targeted investigation.

One human policy decision can authorize a defined proposal category, but agents cannot authorize or apply any correction.

## Current scope

The fresh all-domain queue contains `r scales::comma(n_candidates)` candidates.

It is five candidates larger than the earlier 4,295-row queue because objective invalid-value groups are now included even when within-taxon IQR screening does not flag them.

The workflow generated `r scales::comma(n_proposals)` unapproved proposals: `r scales::comma(n_none_proposals)` conservative no-action proposals and `r scales::comma(n_exclusion_proposals)` exact-zero exclusion proposals.

The remaining `r scales::comma(n_investigation)` candidates require targeted investigation; the bounded first queue contains `r scales::comma(n_agent_queue)`, leaving `r scales::comma(n_investigation_deferred)` for later batches.

```{r}
data_candidate_counts <-
  data_candidates |>
  dplyr::count(
    .data[["trait_domain_name"]],
    name = "Candidates"
  ) |>
  dplyr::rename(`Trait domain` = "trait_domain_name")

knitr::kable(data_candidate_counts)
```

## Proposed policy decisions

An `action = "none"` proposal means that the values remain unchanged and the original flag remains auditable.

An exclusion proposal has the exact selector `trait_value <= 0`; it does not remove positive outliers or an entire taxon-domain group.

```{r}
data_proposal_counts <-
  data_proposals |>
  dplyr::count(
    .data[["trait_domain_name"]],
    .data[["action"]],
    name = "Candidates"
  ) |>
  dplyr::rename(
    `Trait domain` = "trait_domain_name",
    `Proposed action` = "action"
  )

knitr::kable(data_proposal_counts)
```

### Exact-zero exclusions

The `r scales::comma(n_exclusion_proposals)` exclusion proposals select `r scales::comma(n_records_proposed_excluded)` exact-zero records.

Negative values are deliberately excluded from this policy category because a large source-specific negative pattern may indicate a transformation rather than disposable measurements.

```{r}
data_exclusion_details <-
  data_recommendations |>
  dplyr::filter(
    .data[["deterministic_outcome"]] == "exclude_invalid_values"
  ) |>
  dplyr::select(
    `Trait domain` = "trait_domain_name",
    Taxon = "taxon_name",
    Records = "n_records_current",
    `Zero records` = "n_zero",
    Datasets = "n_datasets_current",
    `Impact tier` = "candidate_impact_tier"
  ) |>
  dplyr::arrange(.data[["Trait domain"]], .data[["Taxon"]])

knitr::kable(data_exclusion_details)
```

## Before-and-after scenario

The scenario below applies only the proposed exact-zero exclusions; all other records remain unchanged.

```{r}
data_distribution_table <-
  data_distribution_scenarios |>
  dplyr::select(
    `Trait domain` = "trait_domain_name",
    `Records before` = "n_records_before",
    `Proposed exclusions` = "n_records_proposed_excluded",
    `Records after` = "n_records_after",
    `Minimum before` = "minimum_before",
    `Minimum after` = "minimum_after",
    `Median before` = "median_before",
    `Median after` = "median_after",
    `Maximum before` = "maximum_before",
    `Maximum after` = "maximum_after"
  )

knitr::kable(data_distribution_table, digits = 4)
```

## Targeted investigation

Submitted concerns and recovered correction selectors are never converted automatically to no-action decisions.

Negative or non-finite values also require investigation because they can identify a source-level transformation or ingestion problem.

```{r}
data_investigation_counts <-
  data_recommendations |>
  dplyr::filter(
    .data[["acceptance_eligibility"]] == "requires_agent_review"
  ) |>
  dplyr::count(
    .data[["deterministic_outcome"]],
    name = "Candidates"
  ) |>
  dplyr::rename(`Investigation reason` = "deterministic_outcome") |>
  dplyr::arrange(dplyr::desc(.data[["Candidates"]]))

knitr::kable(data_investigation_counts)
```

The first agent queue is capped at 25 high-priority candidates per domain and prioritizes invalid-value patterns, recovered selectors, and unresolved submitted concerns before generic historical drift.

```{r}
data_agent_queue_counts <-
  data_agent_queue |>
  dplyr::count(
    .data[["trait_domain_name"]],
    .data[["agent_batch_id"]],
    name = "Candidates"
  ) |>
  dplyr::rename(
    `Trait domain` = "trait_domain_name",
    `Agent batch` = "agent_batch_id"
  )

knitr::kable(data_agent_queue_counts)
```

## Approval boundary

Approval should be recorded by policy category, with a named human reviewer and review date, only after this report is accepted.

The recommended first approval decision is whether to accept the conservative `none` category and the exact-zero exclusion category separately.

Targeted investigation results remain proposed until a later evidence report supports a human decision.

After raw decisions are approved and applied, the classified-stage queue must be regenerated and reviewed independently.
