---
title: "Focused trait source-unit investigation"
date: "2026-08-31"
format: gfm
---

```{r setup, include=FALSE}
Sys.setenv(R_CONFIG_ACTIVE = "project_traits_reference")
source(here::here("R/___setup_project___.R"))

path_output_directory <-
  here::here(
    "Outputs/Reports/Trait_corrections/source_unit_investigation"
  )
path_trait_store <-
  resolve_pipeline_store_path(
    pipeline_script = "pipeline_traits_reference.R"
  )
vec_domains_focus <-
  base::c(
    "Stem specific density",
    "Leaf mass per area",
    "Leaf nitrogen content per unit mass"
  )
data_unit_contract <-
  tibble::tribble(
    ~trait_domain_name, ~intended_unit, ~provenance_status,
    "Leaf mass per area", "g/m2", "TRY Global Spectrum release",
    "Leaf nitrogen content per unit mass", "mg/g",
    "TRY Global Spectrum release",
    "Stem specific density", "mg/mm3", "TRY Global Spectrum release"
  )
ssd_plausible_minimum <- 0.1
ssd_plausible_maximum <- 1.5
data_traits <-
  targets::tar_read_raw(
    name = "data_traits_source_scaled",
    store = path_trait_store
  ) |>
  dplyr::filter(
    .data[["trait_domain_name"]] %in% vec_domains_focus,
    base::is.finite(.data[["trait_value"]]),
    .data[["trait_value"]] > 0
  )

connection_vegvault <-
  DBI::dbConnect(
    RSQLite::SQLite(),
    here::here("Data/Input/VegVault.sqlite")
  )
data_source_descriptions <-
  DBI::dbGetQuery(
    connection_vegvault,
    "SELECT data_source_id, data_source_desc FROM DatasetSourcesID"
  ) |>
  tibble::as_tibble()
DBI::dbDisconnect(connection_vegvault)

data_source_taxon_medians <-
  data_traits |>
  dplyr::group_by(
    .data[["taxon_name"]],
    .data[["trait_domain_name"]],
    .data[["data_source_id"]]
  ) |>
  dplyr::summarise(
    source_taxon_median = stats::median(.data[["trait_value"]]),
    n_source_taxon_records = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::group_by(
    .data[["taxon_name"]],
    .data[["trait_domain_name"]]
  ) |>
  dplyr::mutate(
    n_sources_for_taxon = dplyr::n(),
    other_source_median = purrr::map_dbl(
      base::seq_len(dplyr::n()),
      function(index_source) {
        if (dplyr::n() == 1L) {
          return(NA_real_)
        }
        stats::median(
          .data[["source_taxon_median"]][-index_source]
        )
      }
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    comparative_ratio =
      .data[["other_source_median"]] /
        .data[["source_taxon_median"]]
  )

data_source_record_summary <-
  data_traits |>
  dplyr::group_by(
    .data[["data_source_id"]],
    .data[["trait_domain_name"]]
  ) |>
  dplyr::summarise(
    n_records = dplyr::n(),
    n_taxa = dplyr::n_distinct(.data[["taxon_name"]]),
    n_datasets = dplyr::n_distinct(.data[["dataset_id"]]),
    n_trait_names = dplyr::n_distinct(.data[["trait_name"]]),
    value_minimum = base::min(.data[["trait_value"]]),
    value_lwr_25 = stats::quantile(.data[["trait_value"]], 0.25),
    value_median = stats::median(.data[["trait_value"]]),
    value_upr_75 = stats::quantile(.data[["trait_value"]], 0.75),
    value_maximum = base::max(.data[["trait_value"]]),
    .groups = "drop"
  )

data_source_comparison <-
  data_source_taxon_medians |>
  dplyr::filter(base::is.finite(.data[["comparative_ratio"]])) |>
  dplyr::group_by(
    .data[["data_source_id"]],
    .data[["trait_domain_name"]]
  ) |>
  dplyr::summarise(
    n_shared_taxa = dplyr::n(),
    ratio_lwr_25 = stats::quantile(.data[["comparative_ratio"]], 0.25),
    ratio_median = stats::median(.data[["comparative_ratio"]]),
    ratio_upr_75 = stats::quantile(.data[["comparative_ratio"]], 0.75),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    nearest_power_of_ten_ratio =
      10^base::round(base::log10(.data[["ratio_median"]])),
    factor_relative_error = base::abs(
      .data[["ratio_median"]] -
        .data[["nearest_power_of_ten_ratio"]]
    ) / .data[["nearest_power_of_ten_ratio"]]
  )

data_factor_support <-
  data_source_taxon_medians |>
  dplyr::filter(base::is.finite(.data[["comparative_ratio"]])) |>
  dplyr::inner_join(
    data_source_comparison |>
      dplyr::select(
        "data_source_id",
        "trait_domain_name",
        "nearest_power_of_ten_ratio"
      ),
    by = dplyr::join_by(data_source_id, trait_domain_name),
    relationship = "many-to-one"
  ) |>
  dplyr::group_by(
    .data[["data_source_id"]],
    .data[["trait_domain_name"]]
  ) |>
  dplyr::summarise(
    fraction_within_factor_two = base::mean(
      .data[["comparative_ratio"]] >=
        .data[["nearest_power_of_ten_ratio"]] / 2 &
        .data[["comparative_ratio"]] <=
          .data[["nearest_power_of_ten_ratio"]] * 2
    ),
    fraction_within_25_percent = base::mean(
      base::abs(
        .data[["comparative_ratio"]] -
          .data[["nearest_power_of_ten_ratio"]]
      ) / .data[["nearest_power_of_ten_ratio"]] <= 0.25
    ),
    .groups = "drop"
  )

data_source_evidence <-
  data_source_record_summary |>
  dplyr::left_join(
    data_source_comparison,
    by = dplyr::join_by(data_source_id, trait_domain_name),
    relationship = "one-to-one"
  ) |>
  dplyr::left_join(
    data_factor_support,
    by = dplyr::join_by(data_source_id, trait_domain_name),
    relationship = "one-to-one"
  ) |>
  dplyr::left_join(
    data_source_descriptions,
    by = dplyr::join_by(data_source_id),
    relationship = "many-to-one"
  ) |>
  dplyr::mutate(
    absolute_log10_ratio = base::abs(
      base::log10(.data[["ratio_median"]])
    ),
    scale_cluster_strength = dplyr::case_when(
      .data[["n_shared_taxa"]] >= 5 &
        .data[["absolute_log10_ratio"]] >= 0.75 &
        .data[["fraction_within_factor_two"]] >= 0.5 ~ "strong",
      .data[["n_shared_taxa"]] >= 2 &
        .data[["absolute_log10_ratio"]] >= 0.75 ~ "investigate",
      TRUE ~ "none"
    )
  )

vec_forced_focus_sources <- base::c(123L, 137L, 211L, 660L)
data_metadata_candidates <-
  data_source_evidence |>
  dplyr::filter(
    .data[["scale_cluster_strength"]] != "none" |
      .data[["data_source_id"]] %in% vec_forced_focus_sources
  ) |>
  dplyr::arrange(
    factor(
      .data[["scale_cluster_strength"]],
      c("strong", "investigate", "none")
    ),
    .data[["trait_domain_name"]],
    dplyr::desc(.data[["n_shared_taxa"]])
  )

data_partition_summary <-
  data_traits |>
  dplyr::semi_join(
    data_metadata_candidates,
    by = dplyr::join_by(data_source_id, trait_domain_name)
  ) |>
  dplyr::group_by(
    .data[["data_source_id"]],
    .data[["trait_domain_name"]],
    .data[["trait_name"]],
    .data[["dataset_id"]],
    .data[["dataset_name"]]
  ) |>
  dplyr::summarise(
    n_records = dplyr::n(),
    n_taxa = dplyr::n_distinct(.data[["taxon_name"]]),
    value_minimum = base::min(.data[["trait_value"]]),
    value_median = stats::median(.data[["trait_value"]]),
    value_maximum = base::max(.data[["trait_value"]]),
    .groups = "drop"
  )

data_priority_candidates <-
  data_metadata_candidates |>
  dplyr::filter(
    .data[["scale_cluster_strength"]] == "strong" |
      .data[["data_source_id"]] %in% vec_forced_focus_sources
  ) |>
  dplyr::select(
    "data_source_id",
    "data_source_desc",
    "trait_domain_name",
    "n_records",
    "n_taxa",
    "n_shared_taxa",
    "value_median",
    "ratio_median",
    "nearest_power_of_ten_ratio",
    "fraction_within_factor_two",
    "scale_cluster_strength"
  )

data_ssd_assessment <-
  data_traits |>
  dplyr::filter(
    .data[["trait_domain_name"]] == "Stem specific density",
    .data[["data_source_id"]] %in% base::c(123L, 137L, 660L)
  ) |>
  dplyr::mutate(
    diagnostic_factor = dplyr::case_when(
      .data[["data_source_id"]] == 123L ~ 0.001,
      .data[["data_source_id"]] == 137L ~ 100,
      .data[["data_source_id"]] == 660L ~ 1000
    ),
    diagnostic_value = .data[["trait_value"]] *
      .data[["diagnostic_factor"]],
    within_plausibility_band =
      .data[["diagnostic_value"]] >= ssd_plausible_minimum &
        .data[["diagnostic_value"]] <= ssd_plausible_maximum
  ) |>
  dplyr::group_by(.data[["data_source_id"]]) |>
  dplyr::summarise(
    n_records = dplyr::n(),
    value_minimum = base::min(.data[["trait_value"]]),
    value_median = stats::median(.data[["trait_value"]]),
    value_maximum = base::max(.data[["trait_value"]]),
    diagnostic_factor = dplyr::first(.data[["diagnostic_factor"]]),
    diagnostic_minimum = base::min(.data[["diagnostic_value"]]),
    diagnostic_median = stats::median(.data[["diagnostic_value"]]),
    diagnostic_maximum = base::max(.data[["diagnostic_value"]]),
    n_within_band = base::sum(.data[["within_plausibility_band"]]),
    n_outside_band = base::sum(!.data[["within_plausibility_band"]]),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    data_source_descriptions,
    by = dplyr::join_by(data_source_id),
    relationship = "many-to-one"
  ) |>
  dplyr::mutate(
    interpretation = dplyr::case_when(
      .data[["data_source_id"]] == 123L ~
        "near-ready; verify that the source values are kg/m3",
      .data[["data_source_id"]] == 660L ~
        "near-ready; verify the source mass-volume unit",
      TRUE ~
        "mixed source; factor 100 leaves residual records outside the band"
    )
  )

data_ssd_137_residual_summary <-
  data_traits |>
  dplyr::filter(
    .data[["trait_domain_name"]] == "Stem specific density",
    .data[["data_source_id"]] == 137L
  ) |>
  dplyr::mutate(
    diagnostic_value = .data[["trait_value"]] * 100,
    residual_class = dplyr::case_when(
      .data[["diagnostic_value"]] < ssd_plausible_minimum ~
        "below plausibility band",
      .data[["diagnostic_value"]] > ssd_plausible_maximum ~
        "above plausibility band",
      TRUE ~ "within plausibility band"
    )
  ) |>
  dplyr::count(
    .data[["residual_class"]],
    name = "n_records"
  )

readr::write_csv(
  data_source_evidence,
  base::file.path(path_output_directory, "trait_source_unit_evidence.csv")
)
readr::write_csv(
  data_metadata_candidates,
  base::file.path(
    path_output_directory,
    "trait_source_unit_metadata_candidates.csv"
  )
)
readr::write_csv(
  data_partition_summary,
  base::file.path(path_output_directory, "trait_source_unit_partitions.csv")
)
readr::write_csv(
  data_ssd_assessment,
  base::file.path(
    path_output_directory,
    "trait_source_unit_ssd_assessment.csv"
  )
)
readr::write_csv(
  data_ssd_137_residual_summary,
  base::file.path(
    path_output_directory,
    "trait_source_unit_ssd_137_residual_summary.csv"
  )
)
```

# Scope and decision status

This is a diagnostic investigation of Stem specific density, Leaf mass per area and Leaf nitrogen content per unit mass after the ten already approved source rules have been applied.
It does not approve or apply any additional correction.

The comparison uses taxa observed in more than one source.
For each source-domain combination, its taxon medians are compared with the median from all other sources for the same taxa.
The nearest power-of-ten ratio describes a comparative scale cluster only; it is not a proposed conversion factor.
Biological differences among sources can create the same pattern.

# Intended unit contract

The intended units are anchored to the [official TRY Global Spectrum release](https://www.try-db.org/TryWeb/Data.php#81), whose six fields match this project's six trait domains.

```{r unit-contract-table, echo=FALSE, results='asis'}
knitr::kable(data_unit_contract)
```

VegVault 1.0.0 did not retain a usable unit field for these records.
Its archived TRY preparation selected `OrigValueStr` rather than TRY's standardized numeric value, and it reciprocated original SLA values without first normalizing the SLA unit.
Therefore source metadata, not cross-source magnitude alone, must establish any conversion.

# Metadata-verification candidates

```{r priority-table, echo=FALSE, results='asis'}
knitr::kable(data_priority_candidates, digits = 5)
```

This table is deliberately a queue for metadata verification.
Rows marked `strong` have a repeatable cross-source scale displacement, but no row is an approved rule.

# Stem specific density

The diagnostic factors below were tested against a broad `r ssd_plausible_minimum`-`r ssd_plausible_maximum` mg/mm3 screening band.
This band is a triage device, not a replacement for source metadata.

```{r density-assessment, echo=FALSE, results='asis'}
knitr::kable(data_ssd_assessment, digits = 5)
```

Sources 123 and 660 are near-ready for source-rule drafting because their entire record sets land inside the screening band after one dimensionally plausible conversion.
They still require confirmation of the original source unit before approval.

Source 137 is not ready for a blanket rule.
Multiplication by 100 leaves residual records outside the screening band, as shown below.

```{r density-137-residual-table, echo=FALSE, results='asis'}
knitr::kable(data_ssd_137_residual_summary)
```

# Leaf mass per area

The intended LMA unit is g/m2.
The source clusters cannot be translated directly into correction factors because some values are original LMA while others were created by reciprocating SLA recorded in unknown original units.
For example, a value near 100 can be valid LMA in g/m2, whereas a value near 0.05 can be a reciprocal expressed as kg/m2 and would require a different conversion.
A log transformation cannot repair this provenance problem.

```{r lma-candidates, echo=FALSE, results='asis'}
knitr::kable(
  data_metadata_candidates |>
    dplyr::filter(
      .data[["trait_domain_name"]] == "Leaf mass per area"
    ) |>
    utils::head(25),
  digits = 5
)
```

# Leaf nitrogen content per unit mass

The intended nitrogen unit is mg/g.
The observed source medians contain clusters compatible with original g/g, percent and mg/g reporting.
Those conventions would imply different conversions, but the current database does not say which unit applies to each record.
The table therefore prioritizes metadata recovery and does not authorize magnitude-based scaling.

```{r nitrogen-candidates, echo=FALSE, results='asis'}
knitr::kable(
  data_metadata_candidates |>
    dplyr::filter(
      .data[["trait_domain_name"]] ==
        "Leaf nitrogen content per unit mass"
    ) |>
    utils::head(25),
  digits = 5
)
```

# Next gate

The immediate next action is source-metadata recovery, preferably from original TRY unit fields or archived source files.
Sources 123 and 660 should be checked first because a single confirmed unit would settle every affected SSD record.
Source 137 needs dataset-level metadata and residual handling rather than a blanket source rule.
LMA and nitrogen require a programmatic source-to-original-unit mapping; cross-source ratios alone are insufficient.

Until that mapping exists, the current six-trait table remains suitable for diagnostic and sensitivity runs, but not for a definitive all-six-trait reanalysis.
No new correction has been applied by this investigation.
