---
format: 
  revealjs:
    slide-number: true
    theme: [default, custom_theme.scss]
    preview-links: auto
    output-file: "presentation_02"
    output-ext: "html"
    logo: "../../../../Documentation/Materials/Logos/BIODYNAMICS.png"
    code-line-numbers: false
    code-overflow: scroll
    code-link: false
    embed-resources: true
    center-title-slide: false
    include-in-header: "fonts-include.html"
    pdf-separate-fragments: false
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
#| include: false
#| eval: true
library(here)

here::i_am(
  "Documentation/Reports/Second_report/Second_presentation/_presentation_02.qmd"
)

source(
  here::here("R/___setup_project___.R")
)

library(jsonlite)
library(qrcode)
library(bipartite)
library(waffle)

path_to_materials <-
  here::here("Documentation/Materials")

path_figures <-
  here::here(path_to_materials, "Figures")

base::dir.create(
  path = path_figures,
  recursive = TRUE,
  showWarnings = FALSE
)

include_local_figure <- function(data_source) {
  knitr::include_graphics(
    path = here::here(
      path_to_materials,
      data_source
    ),
    error = TRUE
  )
}

sel_project <- "project_europe"

Sys.setenv(R_CONFIG_ACTIVE = sel_project)

vec_pipelines <- "pipeline_basic"

set_store <-
  here::here(
    base::paste0(
      get_active_config("target_store"),
      "/",
      vec_pipelines,
      "/"
    )
  )

verbose <- FALSE

check_presence_of_vegvault()
```

```{r}
#| label: dynamic-theme
#| include: false
#| echo: false
#| eval: true
#| message: false

# Colors
colors <-
  jsonlite::fromJSON(
    here::here(
      "Documentation/Reports/Second_report/Second_presentation/colors.json"
    )
  )

colors_definition <-
  purrr::imap(
    .x = colors,
    .f = ~ {
      paste0(
        "$", .y, ": ", unname(.x), ";", "\n",
        ".reveal .bg-", .y, " { background-color: ", unname(.x), "; }\n",
        ".text-color-", .y, " { color: ", unname(.x), " !important ; }\n",
        ".text-background-", .y, " {\n",
        "background-color: ", unname(.x), ";\n",
        "padding: $smallMargin;\n",
        "border-radius: 5px;\n",
        " }", "\n"
      )
    }
  ) %>%
  paste(collapse = "\n")

writeLines(
  text = c(
    "// This file is auto-generated from colors.json. Do not edit directly.",
    colors_definition
  ),
  con = here::here(
    "Documentation/Reports/Second_report/Second_presentation/_colors.scss"
  )
)

# Fonts

fonts <-
  jsonlite::fromJSON(
    here::here(
      "Documentation/Reports/Second_report/Second_presentation/fonts.json"
    )
  )

fonts_definition <-
  c(
    paste0('$mainFont: "', fonts$body, '", "Arial", sans-serif !default;\n'),
    paste0('$headingFont: "', fonts$heading, '", "Courier New", monospace !default;\n'),
    paste0(
      ".text-font-body { font-family: $mainFont; }\n",
      ".text-font-heading { font-family: $headingFont; }\n"
    ),
    paste0(
      "/* Debug font loading - this will help us see if the font is loaded */\n",
      "@supports (font-family: \"", fonts$heading, "\") {\n",
      "  .debug-font-loaded::before {\n",
      "    content: \"", fonts$heading, " font is supported\";\n",
      "    display: block;\n",
      "    font-size: 12px;\n",
      "    color: green;\n",
      "  }\n",
      "}\n\n",
      "/* Force font loading for debugging */\n",
      ".force-press-start {\n",
      "  font-family: \"", fonts$heading, "\", monospace !important;\n",
      "  font-display: swap;\n",
      "}\n"
    )
  )

writeLines(
  text = c(
    "// This file is auto-generated from fonts.json. Do not edit directly.",
    fonts_definition
  ),
  con = here::here(
    "Documentation/Reports/Second_report/Second_presentation/_fonts.scss"
  )
)
```

# {.bg-green .text-color-white}

<br>

:::: {.columns}

::: {.column width="37%"}
:::

::: {.column width="25%"}

```{r}
#| label: biodynamics-logo-0
include_local_figure("Logos/BIODYNAMICS.png")
```

:::

::: {.column width="37%"}
:::

::::

<br>

:::: {.r-hstack}
::: {data-id="box1" style="background: `r colors$white`; width: 120px; height: 15px; margin: 0px;"}
:::

::: {data-id="box2" style="background: `r colors$pink`; width: 550px; height: 15px; margin: 0px;"}
:::

::: {data-id="box3" style="background: `r colors$black`; width: 120px; height: 15px; margin: 0px;"}
:::

::: {data-id="box4" style="background: `r colors$blue`; width: 120px; height: 15px; margin: 0px;"}
:::
::::

::: {.text-color-white .text-font-heading .text-size-heading3}
[BIODYNAMICS]{.text-bold .text-color-black}: Spatio-temporal dynamics of biotic interactions
:::

<br>

:::: {.columns}

::: {.column width="40%"}
:::

::: {.column width="60%" .text-right .text-color-black .text-size-body}
[👤](https://ondrejmottl.github.io/)[Ondřej Mottl]{.text-color-white}

[10.03.2026]{.text-color-black}

:::

::::

## BIODYNAMICS

- 3-year project (2023–2025) \[extended to 2026 due to parental leave\]
- Funded by Czech Science Foundation (GAČR)

### Objectives

::: {.incremental}

1. [Assemble an **interdisciplinary database** that integrates present and past vegetation records with their functional traits and abiotic environments since the Last Glacial Maximum]{.text-color-blue} (✅)
2. Identify how [**biotic interactions**]{.text-background-pink .text-bold} scale up to form and **maintain biodiversity patterns** at various [spatial/temporal/taxonomic]{.text-background-blue .text-bold} scales
3. [Improve the capacity to **predict global biodiversity patterns** by including scale-dependent information about biotic interactions]{.text-color-blue} (maybe❔)

:::

# Biotic interactions

<br>

::: {.incremental}

- [Non-random co-occurrences]{.text-background-blue .text-bold} after accounting for environmental filtering
- Can be positive (**facilitation**) or negative (**competition**)
- Are [scale dependent]{.text-background-green .text-bold} (?)

:::

## Scales (3 axes)

<br>

::: {.incremental}

- **Spatial**
  - _"Biotic interactions from individual sites (local) to whole continents (global)"_
- **Temporal**
  - _"currently observed changes in biodiversity began millennia ago"_
- **Taxonomic**
  - individual taxa
  - phylogenetically cofound groups
  - groups occupying different ecological niches (i.e. Functional Types)

:::


## Research questions

<br>

1. Is there a [**scale dependence**]{.text-background-blue .text-bold} in the  amount of unexplained variation (potentially due to biotic interactions) structuring vegetation?

::: {.incremental}

 - Expectation 1 - Strength of biotic interactions decrease with increasing [spatial scale]{.text-background-green .text-bold}
 - Expectation 2 - The importance of biotic interaction increase in the periods of large changes (climatic - LGM, anthropogenic - last centuries)
 - Expectation 3 - low taxonomic resolution show stronger biotic interactions at larger spatial scales (interaction between spatial and taxonomic scale)

:::

## Research questions

<br>

2. Can we improve the [**predictive capacity**]{.text-background-pink .text-bold} of biodiversity models by including biotic interactions and historical data?

::: {.incremental}

 - Expectation 4 - Including [historical data]{.text-background-brown .text-bold} will improve model predictive capacity, especially at large scales
 - Expectation 5 - Including biotic interactions will improve model predictive capacity, especially at local scales

:::

# VegVault

:::::: {.columns}

::::: {.column width="50%"}

```{r}
#| label: VegVault-logo
include_local_figure("Logos/VegVault_logo.png")
```

:::::

::::: {.column width="50%"}

* A global SQLite database linking [interdisciplinary]{.text-background-green .text-bold} **plot-based** vegetation data! 
* Links to functional **traits**, **climate** variables, and soil properties 
* Built for exploring **biodiversity** dynamics across [time & space]{.text-background-blue .text-bold}
* Open Source & [Reproducible]{.text-background-brown .text-bold}
* 🔗Learn more: [bit.ly/VegVault](https://bit.ly/VegVault)

:::: {.columns}

::: {.column width="30%"}
:::

::: {.column width="40%"}

```{r}
#| label: vegvault-qr-code-generate
#| echo: false
#| message: false
#| warning: false
#| include: false
qr_vegvault <-
  qrcode::qr_code(
    "https://bit.ly/VegVault",
    ecl = "H"
  )

qrcode::generate_svg(
  qrcode = qr_vegvault,
  file = here::here(
    path_to_materials,
    "QR/qr_vegvault.svg"
  ),
  foreground = colors$black,
  background = colors$white
)
```

```{r}
#| label: vegvault-qr-code-display
#| echo: false
include_local_figure("QR/qr_vegvault.svg")
```

:::

::: {.column width="30%"}
:::

::::

:::::

::::::


## VegVault - Overview

~110 GB | 32 tables & 87 variables | 480,000+ Datasets | 13M+ Samples  | 110,000+ Taxa | 11M+ Trait values | 8 Abiotic variables

:::: {.columns}

#### Data sources 

::: {.column width="40%"}

- 🌿 [**Paleoecological records**]{.text-background-brown .text-bold} ([Neotoma]{.text-italic}, [FOSSILPOL]{.text-italic})
- 🌱 [**Contemporary vegetation**]{.text-background-green .text-bold} ([sPlotOpen]{.text-italic}, [BIEN]{.text-italic})
- 🌾 **Functional traits** ([BIEN]{.text-italic}, [TRY]{.text-italic})
- 🌍 **Climate & soil data** ([CHELSA]{.text-italic}, [WoSIS]{.text-italic})  
:::

::: {.column width="60%"}

```{r}
#| label: vegvault-scheme
include_local_figure("VegVault/assembly_viz_04.png")
```
:::

::::


# Methodological framework

<br>

To address the objectives and research questions, I am developing a large code base with [modular]{.text-background-green .text-bold} and [reusable]{.text-background-blue .text-bold} **data-analysis pipelines** 

```{r}
#| label: sjSDM
#| echo: false
#| message: false
#| warning: false
#| include: true
include_local_figure("BIODYNAMICS/sjSDM.png")
```

The main methodology is [Joined Species Distribution Modelling]{.text-background-pink .text-bold} (JSDM), which allows to infer biotic interactions from species cooccurrence patterns after accounting for environmental filtering

::: {.footer}
[{sjSDM}](https://cran.r-project.org/web/packages/sjSDM/vignettes/sjSDM_Introduction.html)
:::

## scaled JSDM - {sjSDM} package

- scales up to large datasets 
- GPU-accelerated via Python / PyTorch

```{r}
include_local_figure("sjSDM/mee313687-fig-0001-m.jpg")
```


::: {.footer}
[Pichler, M., & Hartig, F. (2021)](https://doi.org/10.1111/2041-210X.13687)
:::

# Technical setup {.bg-green}

```{r}
#| label: extract-codecov
#| echo: false
#| message: false
#| warning: false
#| include: false
var_codecov <-
  jsonlite::fromJSON(
    here::here("Documentation/Functions_test_coverage/covr_report_summary.json")
  ) %>%
  purrr::chuck("value")
```

## Technical framework

:::: {.columns}

::: {.column width="25%"}

```{r}
#| label: r-logo
include_local_figure("Logos/R_logo.svg")
```

```{r}
#| label: targets-logo
include_local_figure("Logos/targets_logo.svg")
```

:::

::: {.column width="75%" .incremental}

- [{targets}]{.text-background-blue .text-bold} pipeline management
  - 16 modular _pipe segments_, reusable across pipelines
- [{config}]{.text-background-pink .text-bold} multi-project management
  - one codebase, multiple spatial/temporal/taxonomic configurations
- [{renv}]{.text-background-brown .text-bold} reproducible package environment
- [Functional programming]{.text-background-green .text-bold} — everything is a function
  - `r length(list.files(here::here("R/Functions"), recursive= TRUE))` modular functions in `R/Functions/`
  - each function is [tested and documented]{.text-background-brown .text-bold}
  - current test coverage: `r var_codecov`%

:::

::::

# "Project" + Pipeline

<br>

::: {.incremental}
- To test research questions I develop multiple [data-analysis _pipelines_]{.text-background-blue .text-bold}
   - each pipeline is a sequence of data processing and analysis steps
   - each pipeline is tailored to a specific research question
- Data and configuration are separated into ["project"s]{.text-background-green .text-bold}
   - each project = specific data subset + configuration
   - e.g. different regions, temporal resolution, taxonomic resolutions

:::

## Basic Pipeline

<br>

```{mermaid}
%%| label: mermaid-pipeline-basic-overview
flowchart LR
    Config-.->VegVault
    Config-.->Data_community
    Config-.->Data_abiotic
    Config-.->Data_coords
    Config-.->Model_prep
    Config-.->sjSDM
    Config-.->Model_evaluation
    Config-.->Model_anova
    subgraph Project
        direction TB
        Config
    end
    subgraph Pipeline
        direction LR
        VegVault-->Data_community
        VegVault-->Data_abiotic
        VegVault-->Data_coords
        Data_community-->Alignment
        Data_abiotic-->Alignment
        Data_coords-->Alignment
        Alignment-->Model_prep
        Model_prep-->sjSDM
        sjSDM-->Model_evaluation
        sjSDM-->Model_anova
    end
```

## Basic Pipeline — EU project

### What's included?

:::: {.columns}

::: {.column width="50%"}

**1. Data extraction** (VegVault)

- Europe bbox: lon −10–40°, lat 35–70°
- Time window: 0–20,000 years BP
- Dataset types: fossil pollen archives + modern gridpoints
- 7 bioclimate variables (CHELSA): bio1, bio4, bio6, bio12, bio15, bio18, bio19

:::

::: {.column width="50%"}

**2. Community data**

- Counts → relative pollen %
- Interpolated to 500-year time steps
- Classified to genus-level
- Rare taxa removed (< 1% pollen proportion)
- Sparse taxa removed (< 2 sites or < 10 samples)
- Proportions binarised → presence/absence

:::

::::

## Basic Pipeline — EU project

### What's included?

:::: {.columns}

::: {.column width="50%"}

**3. Abiotic data**

- Interpolated to 500-year time steps
- Collinear predictors removed

**4. Alignment**

- Samples intersected across community, abiotic, and coordinate data
:::

::: {.column width="50%"}
**5. Model preparation**

- Abiotic predictors z-score scaled
- Coordinates projected to metric (ETRS89-LAEA, EPSG:3035)
- 5 spatial Moran Eigenvector Maps (MEV) computed

:::

::::


## Basic Pipeline — EU project

### What's included?

:::: {.columns}

::: {.column width="50%"}
**6. sjSDM fitting**

- Linear environmental formula (age : bioclim)
- Linear spatial (MEV) formula
- GPU-accelerated Bayesian fitting via Python / PyTorch
:::

::: {.column width="50%"}
**7. Evaluation & ANOVA**

- 5-fold cross-validation (AUC)
- Variance partitioning: environment vs. space vs. biotic
:::

::::

## Basic Pipeline — {targets} graph

```{r}
#| label: target-graph-basic
#| echo: false
#| message: false
#| warning: false
#| output: false
#| cache: true
plot_target_graph_basic <-
  targets::tar_visnetwork(
    script = here::here(
      "R/02_Main_analyses",
      paste0(vec_pipelines, ".R")
    ),
    store = set_store,
    targets_only = TRUE,
    outdated = FALSE,
    physics = TRUE,
    level_separation = 150
  )
```

```{r}
#| label: target-graph-basic-plot
#| echo: false
#| message: false
#| warning: false
#| output: true
plot_target_graph_basic
```

# Model results {.bg-green}

## Variance partitioning

```{r}
#| label: variance-partitioning-setup
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true
model_anova <-
  targets::tar_read(
    name = "model_anova",
    store = set_store
  )

# Extract the 7 disjoint fractions (R2 Nagelkerke) and clamp negatives to 0
vec_euler_raw <-
  model_anova |>
  purrr::chuck("results") |>
  dplyr::filter(
    models %in% c("F_A", "F_B", "F_S", "F_AB", "F_AS", "F_BS", "F_ABS")
  ) |>
  dplyr::select(models, `R2 Nagelkerke`) |>
  tibble::deframe()

vec_euler_raw <-
  pmax(vec_euler_raw, 0)

vec_euler <-
  c(
    "Abiotic" = vec_euler_raw[["F_A"]],
    "Associations" = vec_euler_raw[["F_B"]],
    "Spatial" = vec_euler_raw[["F_S"]],
    "Abiotic&Associations" = vec_euler_raw[["F_AB"]],
    "Abiotic&Spatial" = vec_euler_raw[["F_AS"]],
    "Associations&Spatial" = vec_euler_raw[["F_BS"]],
    "Abiotic&Associations&Spatial" = vec_euler_raw[["F_ABS"]]
  )

euler_fit <-
  eulerr::euler(vec_euler)

grDevices::png(
  filename = here::here(path_figures, "variance_partitioning_euler.png"),
  width = 1600,
  height = 900,
  units = "px",
  res = 150
)

plot(
  euler_fit,
  quantities = list(type = "percent"),
  fills = list(
    fill = c("#7FC97F", "#BEAED4", "#FDC086"),
    alpha = 0.5
  ),
  labels = list(font = 2)
)

grDevices::dev.off()
```

```{r}
#| label: variance-partitioning-display
#| echo: false
include_local_figure("Figures/variance_partitioning_euler.png")
```

## Species-to-Species co-occurences

```{r}
#| label: species-associations-setup
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true
mod_jsdm_assoc <-
  targets::tar_read(
    name = "mod_jsdm",
    store = set_store
  )

data_cor_matrix <-
  sjSDM::getCor(mod_jsdm_assoc)

vec_species_order <-
  mod_jsdm_assoc$species

colnames(data_cor_matrix) <- vec_species_order
rownames(data_cor_matrix) <- vec_species_order

# Select top 30 most common species by prevalence in training data
vec_prevalence <-
  base::colSums(mod_jsdm_assoc$data$Y)

vec_top30_species <-
  vec_prevalence |>
  base::sort(decreasing = TRUE) |>
  utils::head(30) |>
  base::names()

data_cor_matrix_sub <-
  data_cor_matrix[vec_top30_species, vec_top30_species]

vec_species_order <-
  vec_top30_species

data_cor_long <-
  tibble::as_tibble(data_cor_matrix_sub, rownames = "species_row") |>
  tidyr::pivot_longer(
    cols = -species_row,
    names_to = "species_col",
    values_to = "correlation"
  ) |>
  dplyr::mutate(
    species_row = base::factor(species_row, levels = vec_species_order),
    species_col = base::factor(species_col, levels = base::rev(vec_species_order))
  )

figure_species_cor <-
  data_cor_long |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = species_row,
      y = species_col,
      fill = correlation
    )
  ) +
  ggplot2::geom_tile() +
  ggplot2::scale_fill_gradient2(
    low = colors$pink,
    mid = colors$white,
    high = colors$blue,
    midpoint = 0,
    limits = c(-1, 1),
    name = "Correlation"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_text(size = 6),
    axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.ticks = ggplot2::element_blank()
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    title = "Species-to-Species associations (Europe)"
  ) +
  ggview::canvas(
    width = 1600,
    height = 1000,
    units = "px"
  )

ggview::save_ggplot(
  plot = figure_species_cor,
  file = here::here(path_figures, "species_cor_matrix.png")
)
```

```{r}
#| label: species-associations-display
#| echo: false
include_local_figure("Figures/species_cor_matrix.png")
```

## Co-occurrences as bipartite networks

```{r}
#| label: community-bipartite-setup
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true
sel_age <- 2000

data_to_fit <-
  targets::tar_read(
    name = "data_to_fit",
    store = set_store
  )

data_community <-
  data_to_fit %>%
  purrr::chuck("data_community_to_fit") %>%
  base::as.data.frame() %>%
  add_dataset_name_column_from_rownames() %>%
  add_age_column_from_rownames() %>%
  tibble::as_tibble()

data_community_timeslice <-
  data_community %>%
  dplyr::filter(age == sel_age) %>%
  dplyr::select(-age) %>%
  tibble::column_to_rownames("dataset_name")

data_community_timeslice_binary <-
  data_community_timeslice

data_community_timeslice_binary[data_community_timeslice_binary > 0] <- 1
```

<br>

Example of time slice `r sel_age` years BP in Europe (each line =  plot-taxon link, taxa on the top, plots on the bottom):

```{r}
#| label: community-bipartite-example-plot-save
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true
grDevices::png(
  filename = here::here(path_figures, "community_bipartite_network.png"),
  width = 1600,
  height = 800,
  units = "px",
  res = 150
)

data_community_timeslice_binary %>%
  as.matrix() %>%
  bipartite::plotweb(
    text_size = 1,
    sorting = "dec",
    empty = TRUE,
    srt = 90,
    lower_labels = FALSE
  )

grDevices::dev.off()
```

```{r}
#| label: community-bipartite-example-plot
#| echo: false
include_local_figure("Figures/community_bipartite_network.png")
```

## Co-occurrences - network metrics through time

<br>

```{r}
#| label: community-bipartite-metrics
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true

vec_ages <-
  data_community %>%
  dplyr::pull(age) %>%
  unique() %>%
  sort() %>%
  rlang::set_names()

list_data_community_binary <-
  purrr::map(
    .progress = verbose,
    .x = vec_ages,
    .f = ~ {
      data_community_timeslice_i <-
        data_community %>%
        dplyr::filter(age == .x) %>%
        dplyr::select(-age) %>%
        tibble::column_to_rownames("dataset_name")

      data_community_timeslice_binary_i <-
        data_community_timeslice_i

      data_community_timeslice_binary_i[data_community_timeslice_binary_i > 0] <- 1

      data_community_timeslice_binary_i %>%
        as.matrix() %>%
        return()
    }
  )

set.seed(900723)
data_network_metrics <-
list_data_community_binary |> 
  purrr::map(
    .progress = verbose,
    .f = ~ bipartite::networklevel(.x, index = c("connectance", "nestedness")) %>%
      as.data.frame() %>%
      rlang::set_names("value") %>%
      tibble::rownames_to_column("metric") %>%
      tidyr::pivot_wider(
        names_from = metric,
        values_from = value
      ) %>%
      janitor::clean_names()
  ) %>%
  dplyr::bind_rows(.id = "age")

figure_bipartite_metrics <-
  data_network_metrics %>%
  dplyr::select(
    age,
    connectance,
    nestedness
  ) %>%
  tidyr::pivot_longer(
    cols = -age,
    names_to = "metric",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    metric_name = dplyr::case_when(
      metric == "connectance" ~ "Connectance",
      metric == "nestedness" ~ "Nestedness",
      TRUE ~ metric
    )
  ) %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = as.numeric(age),
      y = value
    )
  ) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~metric_name, scales = "free_y") +
  ggplot2::theme_minimal() +
  ggplot2::scale_x_continuous(
    transform = "reverse"
  ) +
  ggplot2::labs(
    x = "Age (cal yr BP)",
    y = "Metric value",
    title = "Selected network metrics through time (Europe)"
  ) +
  ggview::canvas(
    width = 1600,
    height = 800,
    units = "px"
  )

ggview::save_ggplot(
  plot = figure_bipartite_metrics,
  file = here::here(path_figures, "community_bipartite_metrics.png")
)
```

```{r}
#| label: community-bipartite-metrics-display
#| echo: false
include_local_figure("Figures/community_bipartite_metrics.png")
```


# Prediction {.bg-green}

```{r}
#| label: eu-prediction-setup
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true

# Read fitted model and associated targets from EU store
mod_jsdm <-
  targets::tar_read(
    name = "mod_jsdm",
    store = set_store
  )

data_coords <-
  targets::tar_read(
    name = "data_coords",
    store = set_store
  )

data_coords_projected <-
  targets::tar_read(
    name = "data_coords_projected",
    store = set_store
  )

model_evaluation <-
  targets::tar_read(
    name = "model_evaluation",
    store = set_store
  )

scale_attributes <-
  data_to_fit |>
  purrr::chuck("scale_attributes")

# EU config values
x_lim <-
  get_active_config(c("vegvault_data", "x_lim"))

y_lim <-
  get_active_config(c("vegvault_data", "y_lim"))

sel_abiotic_var_name <-
  get_active_config(c("vegvault_data", "sel_abiotic_var_name"))

spatial_crs <-
  get_active_config(c("model_fitting", "spatial_crs"))

use_spatial <-
  get_active_config(c("model_fitting", "use_spatial"))

spatial_mode <-
  get_active_config(c("model_fitting", "spatial_mode"))

# Conditionally load spatial targets
data_spatial_mev_core <- NULL
spatial_scale_attributes <- NULL

if (
  isTRUE(use_spatial)
) {
  spatial_scale_attributes <-
    data_to_fit |>
    purrr::pluck("spatial_scale_attributes")

  if (
    spatial_mode == "spatial"
  ) {
    data_spatial_mev_core <-
      targets::tar_read(
        name = "data_spatial_mev_core",
        store = set_store
      )
  }
}

# Presentation-specific settings: coarse grid
sel_grid_resolution <- 1.0

# Taxon with highest AUC for demonstration
# sel_taxa <-
#   model_evaluation |>
#   purrr::chuck("species") |>
#   dplyr::arrange(dplyr::desc(AUC)) |>
#   dplyr::slice(1L) |>
#   dplyr::pull(species)

# select "nice" taxon for visualization (top AUC is "Styrax" but it's a bit too specific for presentation)
sel_taxa <- "Salix"
```

```{r}
#| label: eu-prediction-grid
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true

# Build a regular land-only grid for prediction
data_grid_base <-
  tidyr::expand_grid(
    coord_long = base::seq(
      base::min(x_lim),
      base::max(x_lim),
      by = sel_grid_resolution
    ),
    coord_lat = base::seq(
      base::min(y_lim),
      base::max(y_lim),
      by = sel_grid_resolution
    )
  ) |>
  dplyr::mutate(grid_id = dplyr::row_number())

land_poly <-
  rnaturalearth::ne_countries(
    scale = "medium",
    returnclass = "sf"
  )

data_grid_land_ids <-
  data_grid_base |>
  sf::st_as_sf(
    coords = c("coord_long", "coord_lat"),
    crs = 4326L
  ) |>
  sf::st_filter(land_poly) |>
  sf::st_drop_geometry() |>
  dplyr::pull(grid_id)

data_grid_base <-
  data_grid_base |>
  dplyr::filter(grid_id %in% data_grid_land_ids)

# Project grid coords to metric km (for MEV interpolation)
data_grid_coords_km <-
  data_grid_base |>
  dplyr::mutate(dataset_name = base::paste0("grid_", grid_id)) |>
  dplyr::select(-grid_id) |>
  tibble::column_to_rownames("dataset_name") |>
  project_coords_to_metric(target_crs = spatial_crs)

# Precompute 2-D MEV interpolation to grid (spatial mode only)
data_grid_mev <- NULL

if (
  isTRUE(use_spatial) &&
    spatial_mode == "spatial"
) {
  data_grid_mev <-
    interpolate_mev_to_grid(
      data_coords_projected_train = data_coords_projected,
      data_mev_core = data_spatial_mev_core,
      data_coords_projected_pred = data_grid_coords_km,
      spatial_scale_attributes = spatial_scale_attributes
    )
}
```

```{r}
#| label: eu-prediction-compute
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true

path_chelsa_cache <-
  here::here(base::paste0("Data/Temp/chelsa/", sel_project))

base::dir.create(
  path_chelsa_cache,
  recursive = TRUE,
  showWarnings = FALSE
)

# Download CHELSA rasters and extract climate values at grid points
data_climate <-
  vec_ages |>
  rlang::set_names() |>
  purrr::map(
    .progress = verbose,
    .f = ~ {
      age_i <- .x

      sel_abiotic_var_name |>
        rlang::set_names() |>
        purrr::map(
          .f = ~ {
            rast_bio <-
              get_chelsa_raster(
                chelsa_var = .x,
                age = age_i,
                x_lim = x_lim,
                y_lim = y_lim,
                cache_dir = path_chelsa_cache
              )

            terra::extract(
              rast_bio,
              y = base::cbind(
                data_grid_base$coord_long,
                data_grid_base$coord_lat
              )
            )[[1]]
          }
        ) |>
        tibble::as_tibble() |>
        dplyr::bind_cols(
          data_grid_base |>
            dplyr::select(grid_id, coord_long, coord_lat)
        ) |>
        dplyr::mutate(age = age_i) |>
        tidyr::drop_na()
    }
  ) |>
  purrr::list_rbind()

# Scale abiotic predictors using training scale attributes
data_abiotic_scaled <-
  data_climate |>
  dplyr::select(
    dplyr::any_of(base::names(scale_attributes))
  ) |>
  dplyr::mutate(
    dplyr::across(
      .cols = dplyr::any_of("age"),
      .fns = ~ .x - base::as.numeric(
        scale_attributes[["age"]][["scaled:center"]]
      )
    )
  ) |>
  dplyr::mutate(
    dplyr::across(
      .cols = -dplyr::any_of("age"),
      .fns = ~ {
        col_nm <- dplyr::cur_column()
        center <- base::as.numeric(
          scale_attributes[[col_nm]][["scaled:center"]]
        )
        sc <- base::as.numeric(
          scale_attributes[[col_nm]][["scaled:scale"]]
        )
        (.x - center) / sc
      }
    )
  ) |>
  base::as.data.frame()

# MEV predictors for all grid points (conditional on spatial mode)
mat_sp <- NULL

if (
  isTRUE(use_spatial) &&
    spatial_mode == "spatial" &&
    !base::is.null(data_grid_mev)
) {
  mat_sp <-
    data_grid_mev[
      base::paste0("grid_", data_climate$grid_id), ,
      drop = FALSE
    ] |>
    base::as.data.frame()
}

# Predict occurrence probabilities
mat_pred <-
  sjSDM:::predict.sjSDM(
    mod_jsdm,
    newdata = data_abiotic_scaled,
    SP = mat_sp,
    type = "link"
  )

base::colnames(mat_pred) <- mod_jsdm$species

# Reshape to long format
data_pred_long <-
  dplyr::bind_cols(
    data_climate |>
      dplyr::select(grid_id, coord_long, coord_lat, age),
    tibble::as_tibble(mat_pred)
  ) |>
  tidyr::pivot_longer(
    cols = -c(grid_id, coord_long, coord_lat, age),
    names_to = "taxon",
    values_to = "predicted_value"
  )
```

## 

```{r}
#| label: eu-prediction-plot-save
#| echo: false
#| message: false
#| warning: false
#| include: false
#| cache: true

# Observed in-sample data for overlay
data_observed <-
  data_to_fit |>
  purrr::chuck("data_community_to_fit") |>
  base::as.data.frame() |>
  add_dataset_name_column_from_rownames() |>
  add_age_column_from_rownames() |>
  tibble::as_tibble() |>
  tidyr::pivot_longer(
    cols = -c(dataset_name, age),
    names_to = "taxon",
    values_to = "value"
  ) |>
  dplyr::mutate(
    observed = base::ifelse(value > 0, "present", "absent")
  ) |>
  dplyr::left_join(
    data_coords |>
      tibble::rownames_to_column("dataset_name") |>
      dplyr::select(dataset_name, coord_long, coord_lat),
    by = dplyr::join_by(dataset_name)
  ) |>
  dplyr::filter(taxon == sel_taxa)

figure_eu_prediction <-
  data_pred_long |>
  dplyr::filter(taxon == sel_taxa) |>
  ggplot2::ggplot() +
  ggplot2::facet_wrap(~age, ncol = 15) +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(
      x = coord_long,
      y = coord_lat,
      fill = predicted_value
    ),
    width = sel_grid_resolution,
    height = sel_grid_resolution
  ) +
  ggplot2::geom_path(
    data = ggplot2::map_data("world"),
    mapping = ggplot2::aes(
      x = long,
      y = lat,
      group = group
    ),
    colour = "black",
    linewidth = 0.1,
    inherit.aes = FALSE
  ) +
  ggplot2::geom_point(
    data = data_observed,
    mapping = ggplot2::aes(
      x = coord_long,
      y = coord_lat,
      shape = observed,
      colour = observed
    ),
    size = 0.1
  ) +
  ggplot2::scale_fill_viridis_c(
    option = "plasma",
    name = "Predicted\noccurrence"
  ) +
  ggplot2::scale_shape_manual(
    values = c("present" = 12L, "absent" = 4L),
    name = "Observed"
  ) +
  ggplot2::scale_colour_manual(
    values = c("present" = "red3", "absent" = "black"),
    name = "Observed"
  ) +
  ggplot2::coord_quickmap(
    xlim = x_lim,
    ylim = y_lim
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    plot.title = ggplot2::element_text(hjust = 0.5),
    plot.subtitle = ggplot2::element_text(hjust = 0.5),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank()
  ) +
  ggplot2::labs(
    x = "Longitude",
    y = "Latitude",
    title = base::paste(
      "Predicted occurrence of", sel_taxa, "— Europe"
    ),
    subtitle = base::paste(
      "AUC:",
      model_evaluation |>
        purrr::chuck("species") |>
        dplyr::filter(species == sel_taxa) |>
        dplyr::pull(AUC) |>
        base::round(3)
    )
  ) +
  ggview::canvas(
    width = 3000,
    height = 2000,
    units = "px"
  )

ggview::save_ggplot(
  plot = figure_eu_prediction,
  file = here::here(path_figures, "eu_prediction.png")
)
```

```{r}
#| label: eu-prediction-plot
#| echo: false
#| out-width: "100%"
include_local_figure("Figures/eu_prediction.png")
```

# Open and public {.bg-green}

```{r}
#| label: generate-urls
#| echo: false
#| message: false
#| warning: false
#| include: false
url_gh <-
  base::paste0(
    "https://github.com/OndrejMottl/",
    "BIODYNAMICS-vegetation-cooccurrence"
  )
```

<br>

Everything shown here is fully [open]{.text-background-blue .text-bold} and [reproducible]{.text-background-brown .text-bold}! Code and materials for this presentation are on GitHub:

:::: {.columns}

::: {.column width="70%"}

<br>

::: {.r-fit-text}
[😸 Code on GitHub](`r url_gh`)

[![](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
:::
:::

::: {.column width="30%"}

```{r}
#| label: qr-slides-generate
#| echo: false
#| message: false
#| warning: false
#| include: false
qr_slides <-
  qrcode::qr_code(
    url_gh,
    ecl = "H"
  )

qrcode::generate_svg(
  qrcode = qr_slides,
  file = here::here("Documentation/Materials/QR/qr_presentation02.svg"),
  foreground = colors[["black"]],
  background = colors[["green"]]
)
```

```{r}
#| label: qr-slides-display
#| echo: false
include_local_figure("QR/qr_presentation02.svg")
```

:::

::::
