---
title: "Report to coauthors"
format:
  html:
    theme: default
    mermaid:
      theme: neutral
toc: true
toc-location: right
toc-depth: 3
date: 2025/10/16
date-format: long
date-modified: last-modified
output-file: Report_to_coauthors.html
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
#| include: false
library(here)

here::i_am("Documentation/Reports/First_report/_Report_to_coauthors.qmd")

source(
  here::here("R/___setup_project___.R")
)

library(Hmsc)
library(MCMCvis)
library(tinytable)
library(waffle)
library(bipartite)
```


## Basic Pipeline

```{r}
#| label: setup-pipeline-basic
#| echo: false
#| message: false
#| warning: false
#| include: false
Sys.setenv(R_CONFIG_ACTIVE = "project_cz")

vec_pipelines <- "pipeline_basic"

set_store <-
  vec_pipelines %>%
  paste0(
    get_active_config("target_store"), "/", ., "/"
  ) %>%
  here::here()
```

### Mermaid

```{mermaid}
flowchart LR
    Config-.->VegVault
    Config-.->Classification
    Config-.->Data_interpolated
    Config-.->model
    subgraph Project
        direction TB
        Config
    end
    subgraph Pipeline
        direction LR
        VegVault-->Data_community
        VegVault-->Data_abiotic
        Data_community-->Classification
        Classification-->Data_harmonised
        Data_abiotic-->Data_interpolated
        Data_harmonised-->Data_interpolated
        Data_interpolated-->model
        model-->model_full
        model-->model_null
        model_full-->model_eval
        model_null-->model_eval
        model_eval-->species_associations
        species_associations-->results
    end
```

### Actual graph

```{r}
#| label: target-graph-basic
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Target graph of the basic pipeline"
targets::tar_visnetwork(
  script = paste0(
    "R/02_Main_analyses", "/", vec_pipelines, ".R"
  ) %>%
    here::here(),
  store = set_store,
  targets_only = TRUE,
  outdated = FALSE,
  physics = TRUE,
  level_separation = 150
)

```

### Model validation

```{r}
#| label: model-validation-setup
#| echo: false
#| message: false
#| warning: false
#| include: false
mod_hmsc_with_eval <-
  targets::tar_read(
    name = "mod_hmsc_to_use",
    store = set_store
  )

mod_hmsc <-
  mod_hmsc_with_eval %>%
  purrr::chuck("mod")

n_samples <- get_active_config(
  value = c("model_fitting", "samples")
)

n_validation_folds <- get_active_config(
  value = c("model_fitting", "cross_validation_folds")
)

n_cores <- get_active_config(
  value = c("model_fitting", "n_cores")
)

n_samples_total <- n_samples * n_cores
```

```{r}
#| label: model-validation-Rhat
#| echo: false
#| message: false
#| warning: false
#| include: true
#| fig-cap: "Histogram of R-hat values for the beta parameters of the HMSC model"
mod_beta_summary <-
  Hmsc::convertToCodaObject(mod_hmsc) %>%
  purrr::chuck("Beta") %>%
  MCMCvis::MCMCsummary() %>%
  janitor::clean_names()

mod_beta_summary %>%
  ggplot2::ggplot(
    ggplot2::aes(x = rhat)
  ) +
  ggplot2::geom_histogram(bins = 30) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    x = "R-hat values",
    y = "Count"
  ) +
  ggplot2::geom_vline(xintercept = 1.1, color = "red", linetype = "dashed") +
  ggplot2::xlim(1, NA)
```


We are selecting model based on the WAIC value (there is option to use ThurR2 as well). This is done automatically to selecte NULL vs FULL model.

####  Predictive power for individual species - summary

```{r}
#| label: model-validation-prediction
#| echo: false
#| message: false
#| warning: false
#| include: true

mod_hmsc_with_eval %>%
  purrr::chuck("eval") %>%
  purrr::map(
    .f = ~ .x %>%
      summary()
  ) %>%
  dplyr::bind_rows(.id = "metric") %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    dplyr::across(
      .cols = !dplyr::where(is.character),
      .fns = ~ as.numeric(.x) %>%
        round(3)
    )
  ) %>%
  tinytable::tt()
```

#### variation partitioning

Let's pretend that the FULL model is the one we want to show here.

```{r}
#| label: model-validation-varpart
#| echo: false
#| message: false
#| warning: false
#| include: true

mod_hmsc_full_with_eval <-
  targets::tar_read(
    name = "mod_hmsc_eval_..._full",
    store = set_store
  )
mod_hmsc_full <-
  mod_hmsc_full_with_eval %>%
  purrr::chuck("mod")


mod_hmsc_full %>%
  Hmsc::computeVariancePartitioning() %>%
  purrr::chuck("vals") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("factor") %>%
  tibble::as_tibble() %>%
  tidyr::pivot_longer(
    cols = -factor,
    names_to = "species",
    values_to = "r2"
  ) %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = species,
      y = r2,
      fill = factor
    )
  ) +
  ggplot2::geom_bar(
    stat = "identity",
    position = "fill"
  ) +
  ggplot2::coord_flip(
    ylim = c(0, 1)
  ) +
  ggplot2::labs(
    title = "Variance partitioning by factor",
    subtitle = paste("project:", Sys.getenv("R_CONFIG_ACTIVE")),
    x = "Species",
    y = "Proportion of explained variance"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  ) +
  ggplot2::scale_fill_viridis_d()
```

#### effect sizes of the environmental covariates

```{r}
#| label: model-validation-effect-sizes
#| echo: false
#| message: false
#| warning: false
#| include: true

mod_hmsc_full_coda <-
  Hmsc::convertToCodaObject(mod_hmsc_full)

vec_all_vars <-
  mod_hmsc_full_coda %>%
  purrr::chuck("Beta", 1) %>%
  colnames()

vec_vars_bio1 <-
  vec_all_vars[
    vec_all_vars %>%
      stringr::str_detect("C2") %>%
      which()
  ]

vec_vars_bio12 <-
  vec_all_vars[
    vec_all_vars %>%
      stringr::str_detect("C3") %>%
      which()
  ]

mod_hmsc_full_coda %>%
  purrr::chuck("Beta") %>%
  MCMCplot(
    object = .,
    ci = c(50, 90),
    params = vec_vars_bio1,
    ISB = FALSE,
    ref_ovl = TRUE,
    rank = TRUE,
    main = "Effect sizes for BIO1\n (mean annual temperature)"
  )

mod_hmsc_full_coda %>%
  purrr::chuck("Beta") %>%
  MCMCplot(
    ci = c(50, 90),
    params = vec_vars_bio12,
    ISB = FALSE,
    ref_ovl = TRUE,
    rank = TRUE,
    main = "Effect sizes for BIO12\n (mean annual precipitation)"
  )

```

### Species associations

```{r}
#| label: species-associations-plot
#| echo: false
#| message: false
#| warning: false
#| include: true

data_species_associations_total <-
  targets::tar_read(
    name = "data_species_associations_total",
    store = set_store
  )

data_species_associations_total %>%
  dplyr::filter(
    type == "dataset_name"
  ) %>%
  dplyr::mutate(
    prop_sign_assoc = n_sign_assoc,
    percentage_sign_assoc = round(prop_sign_assoc * 100),
    n_total = 100,
    n_not_sign_assoc = n_total - percentage_sign_assoc
  ) %>%
  dplyr::select(percentage_sign_assoc, n_not_sign_assoc) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "group",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    group_name = dplyr::case_when(
      group == "percentage_sign_assoc" ~ "Significant associations",
      group == "n_not_sign_assoc" ~ "Non-significant associations"
    )
  ) %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(fill = group_name, values = value)
  ) +
  ggplot2::coord_equal() +
  waffle::geom_waffle(
    n_rows = 10,
    size = 0.5,
    colour = "white",
    flip = TRUE,
    make_proportional = TRUE
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      ncol = 1
    )
  ) +
  ggplot2::theme_void() +
  ggplot2::theme(
    legend.position = "bottom"
  )
```

### Prediction on new data

```{r}
#| label: prediction-setup
#| echo: false
#| message: false
#| warning: false
#| include: false
age_lim <-
  get_active_config(
    value = c("vegvault_data", "age_lim")
  )

vec_age_slices <-
  seq(
    min(age_lim),
    max(age_lim),
    get_active_config(
      value = c("data_processing", "time_step")
    )
  )

x_lim <-
  get_active_config(
    value = c("vegvault_data", "x_lim")
  )

x_lim_max <- max(x_lim, na.rm = TRUE)
x_lim_min <- min(x_lim, na.rm = TRUE)

y_lim <-
  get_active_config(
    value = c("vegvault_data", "y_lim")
  )

y_lim_max <- max(y_lim, na.rm = TRUE)
y_lim_min <- min(y_lim, na.rm = TRUE)
```

```{r}
#| label: prediction-new-data
#| echo: false
#| message: false
#| warning: false
#| include: true
#| cache: true
con <-
  DBI::dbConnect(
    RSQLite::SQLite(),
    here::here("Data/Input/VegVault.sqlite")
  )

data_gridpoints_raw <-
  dplyr::tbl(con, "Datasets") %>%
  dplyr::filter(data_source_type_id == 5) %>%
  dplyr::filter(
    coord_long >= x_lim_min &
      coord_long <= x_lim_max &
      coord_lat >= y_lim_min &
      coord_lat <= y_lim_max
  ) %>%
  dplyr::inner_join(
    dplyr::tbl(con, "DatasetSample"),
    by = "dataset_id"
  ) %>%
  dplyr::inner_join(
    dplyr::tbl(con, "Samples"),
    by = "sample_id"
  ) %>%
  dplyr::inner_join(
    dplyr::tbl(con, "AbioticData") %>%
      dplyr::filter(
        abiotic_variable_id == 1 | # BIO1
          abiotic_variable_id == 4 # BIO12
      ),
    by = "sample_id"
  ) %>%
  dplyr::collect()

DBI::dbDisconnect(con)

```

```{r}
#| label: prediction-new-data-wrangle
#| echo: false
#| message: false
#| warning: false
#| include: true

data_to_predict <-
  data_gridpoints_raw %>%
  dplyr::filter(
    age %in% vec_age_slices
  ) %>%
  dplyr::select(
    dataset_name,
    coord_long, coord_lat,
    age,
    abiotic_variable_id, abiotic_value
  )


data_to_predict_distinct <-
  data_to_predict %>%
  tidyr::pivot_wider(
    names_from = abiotic_variable_id,
    values_from = abiotic_value,
    names_prefix = "var_id_"
  ) %>%
  dplyr::distinct(coord_long, coord_lat, age, .keep_all = TRUE) %>%
  dplyr::rename(
    bio1 = var_id_1,
    bio12 = var_id_4
  )

data_to_predict_clim <-
  data_to_predict_distinct %>%
  dplyr::select(bio1, bio12) %>%
  as.data.frame()


data_coords <-
  data_to_predict_distinct %>%
  dplyr::distinct(dataset_name, coord_long, coord_lat) %>%
  tibble::column_to_rownames("dataset_name")

random_coors_knots <-
  Hmsc::constructKnots(
    sData = data_coords,
    minKnotDist = get_active_config(
      c("data_processing", "min_distance_of_gpp_knots")
    )
  )

random_level_coords <-
  Hmsc::HmscRandomLevel(
    sData = data_coords,
    sMethod = "GPP",
    sKnot = random_coors_knots
  )

data_age <-
  data_to_predict_distinct %>%
  dplyr::distinct(age) %>%
  dplyr::arrange(age) %>%
  dplyr::mutate(
    age_rowname = as.character(age)
  ) %>%
  tibble::column_to_rownames("age_rowname")

random_level_age <-
  Hmsc::HmscRandomLevel(
    sData = data_age
  )

study_design <-
  data_to_predict_distinct %>%
  dplyr::mutate(
    age_factor = factor(
      x = age,
      levels = sort(unique(as.numeric(age))),
      ordered = TRUE
    ),
    dataset_name_factor = as.factor(dataset_name)
  ) %>%
  dplyr::select(
    "dataset_name" = "dataset_name_factor",
    "age" = "age_factor"
  ) %>%
  as.data.frame()


```

```{r}
#| label: prediction-new-data-predict
#| echo: false
#| message: false
#| warning: false
#| include: true
#| cache: true
data_pred <-
  Hmsc:::predict.Hmsc(
    object = mod_hmsc_full,
    XData = data_to_predict_clim,
    studyDesign = study_design,
    ranLevels = list(
      age = random_level_age,
      dataset_name = random_level_coords
    ),
    predictEtaMean = TRUE,
    expected = TRUE
  )
```

```{r}
#| label: prediction-new-data-postprocess
#| echo: false
#| message: false
#| warning: false
#| include: true
#| cache: true
mean_with_na <- function(x) {
  mean(x, na.rm = TRUE)
}

data_pred_avg <-
  apply(
    abind::abind(data_pred, along = 3),
    c(1, 2),
    FUN = mean_with_na
  )
```

```{r}
#| label: prediction-new-data-get-observed
#| echo: false
#| message: false
#| warning: false
#| include: true

data_to_fit <-
  targets::tar_read(
    name = "data_to_fit",
    store = set_store
  )

data_taxa_eval <-
  tibble::tibble(
    taxon = data_to_fit %>%
      purrr::chuck("data_community_to_fit") %>%
      colnames(),
    mod_hmsc_full_with_eval %>%
      purrr::chuck("eval") %>%
      as.data.frame()
  )

data_observed <-
  data_to_fit %>%
  purrr::chuck("data_community_to_fit") %>%
  add_age_column_from_rownames() %>%
  add_dataset_name_column_from_rownames() %>%
  as_tibble() %>%
  dplyr::left_join(
    data_to_fit %>%
      purrr::chuck("data_coords_to_fit") %>%
      tibble::rownames_to_column("dataset_name"),
    by = "dataset_name"
  ) %>%
  tidyr::pivot_longer(
    cols = -c(age, dataset_name, coord_long, coord_lat),
    names_to = "taxon",
    values_to = "proportion"
  ) %>%
  dplyr::mutate(
    present = ifelse(proportion > 0, "present", "absent")
  )

data_predicted <-
  as_tibble(study_design) %>%
  dplyr::mutate(
    coord_long = dataset_name %>%
      stringr::str_remove("geo_") %>%
      stringr::str_remove("_.*"),
    coord_lat = dataset_name %>%
      stringr::str_remove("geo_") %>%
      stringr::str_remove(".*_"),
    age = as.numeric(as.character(age))
  ) %>%
  dplyr::bind_cols(
    as_tibble(data_pred_avg)
  ) %>%
  tidyr::pivot_longer(
    cols = -c(dataset_name, coord_lat, coord_long, age),
    names_to = "taxon",
    values_to = "pred_value"
  ) %>%
  dplyr::select(taxon, coord_long, coord_lat, age, pred_value)
```

#### Single species plot (biogeography)

We will fist plot the predicted values for a single species across all time slices.

```{r}
#| label: prediction-new-data-plot-single-species
#| echo: false
#| message: false
#| warning: false
#| include: true
#| out-width: "100%"
#| out-height: "100%"
#| fig-height: 7
sel_taxa <- "Fraxinus"

data_observed_sel_taxa <-
  data_observed %>%
  dplyr::filter(taxon == sel_taxa)

data_predicted_sel_taxa <-
  data_predicted %>%
  dplyr::filter(taxon == sel_taxa)

# Make a plot
data_predicted_sel_taxa %>%
  ggplot2::ggplot() +
  ggplot2::facet_wrap(~age) +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(
      x = as.numeric(coord_long),
      y = as.numeric(coord_lat),
      fill = pred_value
    ),
    col = NA,
    alpha = 0.8,
    linewidth = 0,
    width = 0.5,
    height = 0.5
  ) +
  ggplot2::borders(
    "world",
    fill = NA,
    colour = "black"
  ) +
  ggplot2::geom_point(
    data = data_observed_sel_taxa,
    mapping = ggplot2::aes(
      x = as.numeric(coord_long),
      y = as.numeric(coord_lat),
      color = as.factor(present),
      shape = as.factor(present)
    ),
    size = 5,
    alpha = 0.5
  ) +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::scale_color_manual(
    values = c(
      "present" = "red",
      "absent" = "black"
    )
  ) +
  ggplot2::scale_shape_manual(
    values = c(
      "present" = 12,
      "absent" = 13
    )
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "top",
    plot.title = ggplot2::element_text(hjust = 0.5),
    plot.subtitle = ggplot2::element_text(hjust = 0.5)
  ) +
  ggplot2::coord_quickmap(
    xlim = x_lim,
    ylim = y_lim
  ) +
  ggplot2::labs(
    fill = "Predicted\noccurrence",
    color = "Observed",
    shape = "Observed",
    x = "Longitude",
    y = "Latitude",
    title = paste(
      "Predicted occurrence of",
      sel_taxa,
      get_active_config(
        c("data_processing", "taxonomic_resolution")
      ),
      " for Czechia"
    ),
    subtitle = "Each facet is a different time slice",
    caption = paste(
      paste(
        "HMSC model with space (GPP; minimum distance of",
        get_active_config(
          c("data_processing", "min_distance_of_gpp_knots")
        ),
        ") and age as random effect"
      ),
      paste(
        "Climatic variables:",
        paste(
          get_active_config(
            c("vegvault_data", "sel_abiotic_var_name")
          ),
          collapse = ", "
        )
      ),
      paste(
        "Model fitted with",
        n_samples_total,
        "samples and",
        get_active_config(
          c("model_fitting", "cross_validation_folds")
        ),
        "cross-validation folds"
      ),
      paste(
        "Model evaluation:",
        data_taxa_eval %>%
          dplyr::filter(taxon == sel_taxa) %>%
          dplyr::pull(AUC) %>%
          round(3),
        "AUC;",
        data_taxa_eval %>%
          dplyr::filter(taxon == sel_taxa) %>%
          dplyr::pull(TjurR2) %>%
          round(3),
        "TjurR2;",
        data_taxa_eval %>%
          dplyr::filter(taxon == sel_taxa) %>%
          dplyr::pull(RMSE) %>%
          round(3),
        "RMSE"
      ),
      sep = "\n"
    )
  )

```

#### Diversity through time

```{r}
#| label: prediction-diversity-estimation
#| echo: false
#| message: false
#| warning: false
#| include: false

sel_threshold <- 0.5

data_predicted_richness <-
  seq(0.25, 0.75, by = 0.05) %>%
  rlang::set_names() %>%
  purrr::map(
    .f = ~ data_predicted %>%
      dplyr::mutate(
        present = ifelse(pred_value >= .x, TRUE, FALSE)
      ) %>%
      dplyr::group_by(coord_long, coord_lat, age) %>%
      dplyr::summarise(
        richness = sum(present, na.rm = TRUE),
        .groups = "drop"
      )
  ) %>%
  dplyr::bind_rows(.id = "threshold") %>%
  dplyr::group_by(coord_long, coord_lat, age) %>%
  dplyr::summarise(
    richness_mean = mean(richness, na.rm = TRUE),
    richness_min = min(richness, na.rm = TRUE),
    richness_max = max(richness, na.rm = TRUE),
    richness_range = abs(richness_max - richness_min),
    richness_sd = sd(richness, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: prediction-diversity-plot
#| echo: false
#| message: false
#| warning: false
#| include: true
#| out-width: "100%"
#| out-height: "100%"
#| fig-height: 7

data_predicted_richness %>%
  ggplot2::ggplot() +
  ggplot2::facet_wrap(~age) +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(
      x = as.numeric(coord_long),
      y = as.numeric(coord_lat),
      fill = richness_mean,
      alpha = 1 / richness_range
    ),
    col = NA,
    linewidth = 0,
    width = 0.25,
    height = 0.25
  ) +
  ggplot2::borders(
    "world",
    fill = NA,
    colour = "black"
  ) +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "top",
    plot.title = ggplot2::element_text(hjust = 0.5),
    plot.subtitle = ggplot2::element_text(hjust = 0.5)
  ) +
  ggplot2::coord_quickmap(
    xlim = x_lim,
    ylim = y_lim
  ) +
  ggplot2::labs(
    fill = "Predicted\nrichness",
    x = "Longitude",
    y = "Latitude",
    alpha = "Prediction confidence\n(1/value range)",
    title = paste(
      "Predicted taxonomic richness for Czechia"
    ),
    subtitle = "Each facet is a different time slice",
    caption = paste(
      paste(
        "HMSC model with space (GPP; minimum distance of",
        get_active_config(
          c("data_processing", "min_distance_of_gpp_knots")
        ),
        ") and age as random effect"
      ),
      paste(
        "Climatic variables:",
        paste(
          get_active_config(
            c("vegvault_data", "sel_abiotic_var_name")
          ),
          collapse = ", "
        )
      ),
      paste(
        "Model fitted with",
        get_active_config(
          c("model_fitting", "n_cores")
        ),
        "cores,",
        get_active_config(
          c("model_fitting", "samples")
        ),
        "samples,",
        get_active_config(
          c("model_fitting", "cross_validation_folds")
        ),
        "cross-validation folds"
      ),
      sep = "\n"
    )
  )

```

## Pipeline with time

```{mermaid}
flowchart LR
    Config-.->VegVault
    Config-.->Classification
    Config-.->Data_interpolated
    Config-.->Data_subset
    subgraph Project
        direction TB
        Config
    end
    subgraph Pipeline
        direction LR
        VegVault-->Data_community
        VegVault-->Data_abiotic
        Data_community-->Classification
        Classification-->Data_harmonised
        Data_abiotic-->Data_interpolated
        Data_harmonised-->Data_interpolated
        Data_interpolated-->Data_subset
        species_associations_1-->results
        species_associations_2-->results
        species_associations_3-->results
        subgraph Age_slices
            direction LR
            Data_subset-->model_age_1
            Data_subset-->model_age_2
            Data_subset-->model_age_3
            model_age_1-->species_associations_1
            model_age_2-->species_associations_2
            model_age_3-->species_associations_3
        end
end
```

```{r}
#| label: target-graph-time
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Target graph of the time pipeline"
Sys.setenv(R_CONFIG_ACTIVE = "project_cz")
vec_pipelines <- "pipeline_time"

set_store <-
  vec_pipelines %>%
  paste0(
    get_active_config("target_store"), "/", ., "/"
  ) %>%
  here::here()

targets::tar_visnetwork(
  script = paste0(
    "R/02_Main_analyses", "/", vec_pipelines, ".R"
  ) %>%
    here::here(),
  store = set_store,
  outdated = FALSE,
  targets_only = TRUE,
  physics = TRUE,
  level_separation = 150
)

```

### Coocurences in time slices


```{r}
#| label: cooccurrence-time-slices
#| echo: false
#| message: false
#| warning: false
#| include: true

targets::tar_read(
  name = "plot_species_associations",
  store = set_store
)

```

### Coocurences as food webs

We can think about the co-occurrence networks as bipartite networks (e.g., plants and their pollinators), where individual plots are one layer and species are another layer. 


Example of one time slice:
```{r}
#| label: cooccurrence-bipartite-setup
#| echo: false
#| message: false
#| warning: false
#| include: true
#| cache: true
data_community <-
  data_to_fit %>%
  purrr::chuck("data_community_to_fit") %>%
  add_dataset_name_column_from_rownames() %>%
  add_age_column_from_rownames() %>%
  tibble::as_tibble()

data_community_timeslice <-
  data_community %>%
  dplyr::filter(age == 2000) %>%
  dplyr::select(-age) %>%
  tibble::column_to_rownames("dataset_name")

data_community_timeslice_binary <-
  data_community_timeslice

data_community_timeslice_binary[data_community_timeslice_binary > 0] <- 1

data_community_timeslice_binary %>%
  as.matrix() %>%
  bipartite::plotweb(
    text_size = 1,
    sorting = "dec",
    empty = TRUE,
    srt = 90,
    lower_labels = FALSE
  )

```


We can then calculate network-level metrics for these bipartite networks.

```{r}
#| label: cooccurrence-bipartite
#| echo: false
#| message: false
#| warning: false
#| include: true
#| cache: true

vec_ages <-
  data_community %>%
  dplyr::pull(age) %>%
  unique() %>%
  sort() %>%
  rlang::set_names()

list_data_community_binary <-
  purrr::map(
    .x = vec_ages,
    .f = ~ {
      data_community_timeslice <-
        data_community %>%
        dplyr::filter(age == .x) %>%
        dplyr::select(-age) %>%
        tibble::column_to_rownames("dataset_name")

      data_community_timeslice_binary <-
        data_community_timeslice

      data_community_timeslice_binary[data_community_timeslice_binary > 0] <- 1

      data_community_timeslice_binary %>%
        as.matrix() %>%
        return()
    }
  )

set.seed(1234)
data_network_metrics <-
  purrr::map(
    .progress = TRUE,
    .x = list_data_community_binary,
    .f = ~ bipartite::networklevel(.x) %>%
      as.data.frame() %>%
      rlang::set_names("value") %>%
      tibble::rownames_to_column("metric") %>%
      tidyr::pivot_wider(
        names_from = metric,
        values_from = value
      ) %>%
      janitor::clean_names()
  ) %>%
  dplyr::bind_rows(.id = "age")

```

Here we plot several metrics through time.

```{r}
#| label: cooccurrence-bipartite-plot
#| echo: false
#| message: false
#| warning: false
#| include: true
data_network_metrics %>%
  dplyr::select(
    age,
    connectance,
    nestedness,
    modularity_q
  ) %>%
  tidyr::pivot_longer(
    cols = -age,
    names_to = "metric",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    metric_name = dplyr::case_when(
      metric == "connectance" ~ "Connectance",
      metric == "nestedness" ~ "Nestedness",
      metric == "modularity_q" ~ "Modularity (Q)",
      TRUE ~ metric
    )
  ) %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = as.numeric(age),
      y = value
    )
  ) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~metric_name, scales = "free_y") +
  ggplot2::theme_minimal() +
  ggplot2::scale_x_continuous(
    transform = "reverse"
  ) +
  ggplot2::labs(
    x = "Age (cal yr BP)",
    y = "Metric value",
    title = "Selected metrics through time"
  )
```
